---
format: 
   revealjs:
      embed-resources: true
      theme: serif
      slide-level: 3
      slide-number: true
      toc-depth: 2
      show-slide-number: all
      preview-links: auto
      number-sections: true
      link-color: orange
      include-after-body: menu-hide-level3.html
title: "Lectures on causal inference and experimental methods"
author: "Macartan Humphreys"
bibliography: "../../slides/bib.bib"
---


# Roadmap

1. Intro | [Course outline, tools](#secoutline) |  [`DeclareDesign`](#secdd)
2. Causality | [Fundamental problems and solutions](#seccausality) | [Inquiries and identification](#secestimands) 
3. Estimation and Inference 1:  [Frequentist](#secfisher) 
4. Estimation and Inference 2: [Bayesian](#secbayes)
5. Design |  [Experimental Design](#secdesign) |  [Design evaluation](#secdiagnosis)
6. Topics 1| [Topics and techniques 1](#citopics1) 
7. Topics 2| [Topics and techniques 2](#citopics2) 



```{css}
##| echo: false
##| include: false

.reveal table {
  font-size: medium;
}

/* Left menu: show first and second levels (all items); hide third level */
.slide-menu-wrapper .slide-menu-item[data-level="3"],
.slide-menu-wrapper .slide-menu-item-vertical[data-level="3"] {
  display: none;
}
.slide-menu-wrapper .slide-menu-item-vertical {
  padding-left: 1.5em;
}

## also: smaller, small

```
  
```{r, include = FALSE}

source(here::here("slides", "setup.R"))
options(future.globals.maxSize = 3 * 1024^3)  # 2GB

# remotes::install_github("jhelvy/renderthis")
# renderthis::to_pdf("ci_2024.html")
```

```{r, child = here::here("slides", "1.1_intro.qmd")}

```


# `DeclareDesign` {#secdd}

*How to define and assess research designs* 

```{r, child = here::here("slides","1.2_declaredesign.qmd")}

```


# Causality {#seccausality} 

```{r, child = here::here("slides","2.1_causality.qmd")}

```

# Inquiries {#secestimands}

*Well posed questions*

```{r, child = here::here("slides","2.2_estimands.qmd")}

```

# Frequentist Analysis {#secfisher}

*Estimation and testing*

```{r, child = here::here("slides", "3_fisher.qmd")}

```


```{r, child = here::here("slides", "4_bayes.qmd")}

```

# Design {#secdesign}

```{r, child = here::here("slides", "5.1_design.qmd")}

```

# Design evaluation {#secdesign2}

```{r, child = here::here("slides", "5.2_evaluation.qmd")}

```


```{r top61, child = here::here("slides","6_topics_1.qmd")}

```


```{r, child = here::here("slides","7_topics_2.qmd")}

```


```{r, child = here::here("slides","8_topics_3.qmd")}

```


# Resources

### References
