---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
title: "Causality"
author: "Macartan Humphreys"
bibliography: bib.bib
---

# The idea of counterfactual causation

```{r, include = FALSE}

source("setup.R")

```



```{r pos, child = "2.1.1_potential_outcomes.qmd"}

```

# Estimands 

The inquiries you have...

```{r estms, child = "2.2.1_estimand_types.qmd"}

```


### Recap: Things you need to know about causal inference {.smaller}

1.  A causal claim is a statement about what didn't happen.
4.  If you know that $A$ causes $B$ and that $B$ causes $C$, this does not mean that you know that $A$ causes $C$.
10. There is no causation without manipulation.
2.  There is a fundamental problem of causal inference.
3.  You can estimate average causal effects even if you cannot observe any individual causal effects.
9.  Estimating average causal effects via differences in means does not require that treatment and control groups are identical.
9.  Estimating average causal effects via differences in means is fraught when you condition on post treatment variables or on colliders.


# References
