---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
    link-color: orange
title: "Inquiries and identification"
author: "Macartan Humphreys"
bibliography: bib.bib
---

```{r, include = FALSE}

source(here::here("slides", "setup.R"))

```


## Outline

* Types of estimands
* Principal strata
* Identification
* Backdoor
* Frontdoor
* `dagitty`

```{r, child = here::here("slides", "2.2.1_estimand_types.qmd")}

```


## Inquiries as statements about principal strata

Given a causal model over nodes with discrete ranges, inquiries can generally be described as summaries of the distributions of exogenous nodes.

We already saw two instances of this:

-   The probability that $X$ has a positive effect on $Y$ in an $X \rightarrow Y$ model is $\lambda^Y_{01}$ (last lecture)
-   The share of "compliers" in an IV model $Z \rightarrow X \rightarrow Y \leftrightarrow X$ is $\lambda^X_{01}$

```{r, child = here::here("slides", "2.2_2_identification.qmd")}

```



