---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
    link-color: orange
title: "Inquiries and identification"
author: "Macartan Humphreys"
bibliography: bib.bib
---

```{r, include = FALSE}

source(here::here("slides", "setup.R"))

```




## Identification

*What it is. When you have it. What it's worth.*

### Identification

Informally a quantity is "identified" if it can be "recovered" once you have enough data.

Say for example average wage is $x$ in some very large population. If I gather lots and lots of data on the wages of individuals and take the average then then my estimate will ultimately let be figure out $x$.

-   If $x$ is \$1 then my estimate will end up centered on \$1.
-   If it is \$2 it will end up centered on \$2.

### Identification (Definition) {.smaller}

-   **Identifiability** Let $Q(M)$ be a query defined over a class of models $\mathcal M$, then $Q$ is identifiable if $P(M_1) = P(M_2) \rightarrow Q(M_1) = Q(M_1)$.

-   **Identifiability with constrained data** Let $Q(M)$ be a query defined over a class of models $\mathcal M$, then $Q$ is identifiable from features $F(M)$ if $F(M_1) = F(M_2) \rightarrow Q(M_1) = Q(M_1)$.

Based on Defn 3.2.3 in Pearl.

-   **Essentially**: Each underlying value produces a unique data distribution. When you see that distribution you recover the parameter.

### Example without identification 1 {.smaller}

Informally a quantity is "identified" if it can be "recovered" once you have enough data.

-   Say for example average wage is $x^m$ for men and $x^w$ for women (in some very large population).
-   If I gather lots and lots of data on the wages of (male and female) couples, e.g. $x^c_i = x^m_i + x^w_i$ then, although this will be informative, it will never be sufficient to recover $x^m$ for men and $x^w$.
-   I can recover $x^c$, but there are too many combinations of possible values of $x^m$ and $x^w$ consistent with the observed data.


### Example without identification 2 {.smaller}

* share $a$ of units have  negative effects
* $b$ have positive effects
* $c$ that always have $Y=0$ and 
* $d$ always have  $Y=1$.

Then with *very large* (experimental) data we observe:

|     | Y = 0 | Y = 1 | 
|-----|-------|-------|
| **X = 0** | $\alpha_{00} \rightarrow b/2 + c/2$      |  $\alpha_{01} \rightarrow a/2 + d/2$     |   
| **X = 1** | $\alpha_{10} \rightarrow a/2 + c/2$      |  $\alpha_{11} \rightarrow b/2 + d/2$     |   

What quantities are identified? 

* $b-a$?
* $d-c$?
* $b$?


### Example without identification 2 {.smaller}

What if we :

* *knew* that $a=0$
* *observed* that $\alpha_01=0$

|     | Y = 0 | Y = 1 | 
|-----|-------|-------|
| **X = 0** | $\alpha_{00} \rightarrow b/2 + c/2$      |  $\alpha_{01} \rightarrow a/2 + d/2$     |   
| **X = 1** | $\alpha_{10} \rightarrow a/2 + c/2$      |  $\alpha_{11} \rightarrow b/2 + d/2$     |   

What quantities are now identified? 

* $b-a$?
* $d-c$?
* $b$?


### Identification : Goal {.smaller}

Our goal in causal inference is to estimate quantities such as:

$$\Pr(Y|\hat{x})$$

where $\hat{x}$ is interpreted as $X$ set to $x$ by "external" control. Equivalently: $do(X=x)$ or sometimes $X \leftarrow x$.

-   If this quantity is **identifiable** then we can recover it with infinite data.

-   If it is not identifiable, then, even in the best case, we are not guaranteed to get the right answer.

*Are there general rules for determining whether this quantity can be identified?* Yes.

### Identification : Goal

Note first, identifying

$$\Pr(Y|x)$$

is easy.

But we are not always interested in identifying the distribution of $Y$ given *observed* values of $x$, but rather, the distribution of $Y$ if $X$ is *set* to $x$.

## Levels and effects

If we can identify the controlled distribution we can calculate other causal quantities of interest.

For example for a binary $X, Y$ the causal effect of $X$ on the probability that $Y=1$ is:

$$\Pr(Y=1|\hat{x}=1) - \Pr(Y=1|\hat{x}=0)$$

Again, **this is not the same as**:

$$\Pr(Y=1|x=1) - \Pr(Y=1|x=0)$$

It's the difference between seeing and doing.

### When to condition? What to condition on? {.smaller}

The key idea is that you want to find a set of variables such that when you condition on these you get what you would get if you used a `do` operation.

Intuition:

-   You could imagine creating a "mutilated" graph by removing all the arrows leading *out* of *X*
-   Then select a set of variables, $Z$, such that $X$ and $Y$ are d-separated by $Z$ on the the mutilated graph
-   When you condition on these you are making sure that any covariation between $X$ and $Y$ is covariation that is due to the effects of $X$

### Illustration

```{r, echo = FALSE}
m1 <- make_model("X -> M1 -> Y <- M2 <- X <- Z -> M1")
plot(m1, x_coord = c(1, 2, 3, 3, 4), y_coord = c(1.5, 1, 2, 1, 1))
```

### Illustration: Remove paths out

```{r, echo = FALSE}
m2 <- make_model("M1 -> Y <- M2; X <- Z -> M1")
plot(m2, x_coord = c(M2 = 3, Z=1, M1=3, X=2, Y=4), y_coord = c(M2=1, Z=1.5, M1=2, X=1, Y=1))
```

### Illustration: Block backdoor path

```{r, echo = FALSE}
m2 <- make_model("M1 -> Y <- M2; X <- Z -> M1")
plot(m2, 
     x_coord = c(M2 = 3, Z=1, M1=3, X=2, Y=4), 
     y_coord = c(M2=1, Z=1.5, M1=2, X=1, Y=1), 
     nodecol = c(1,2,1,1,1))
```

### Illustration: Why not like this?

```{r, echo = FALSE}
m2 <- make_model("M1 -> Y <- M2; X <- Z -> M1")
plot(m2, 
     x_coord = c(M2 = 3, Z=1, M1=3, X=2, Y=4), 
     y_coord = c(M2=1, Z=1.5, M1=2, X=1, Y=1), 
     nodecol = c(1,1,2, 1,1))
```

### Identification

-   Three results ("Graphical Identification Criteria")
    -   Backdoor criterion
    -   Adjustment criterion
    -   Frontdoor criterion
-   There are more

### Backdoor Criterion: (Pearl 1995) {.smaller}

The **backdoor criterion** is satisfied by $Z$ (relative to $X$, $Y$) if:

1.  No node in $Z$ is a descendant of $X$
2.  $Z$ blocks every **backdoor** path from $X$ to $Y$ (i.e. every path that contains an arrow into $X$)

In that case you can identify the effect of $X$ on $Y$ by conditioning on $Z$:

$$P(Y=y | \hat{x}) = \sum_z P(Y=y| X = x, Z=z)P(z)$$ (This is eqn 3.19 in Pearl (2000))

### Backdoor Criterion: (Pearl 1995)

$$P(Y=y | \hat{x}) = \sum_z P(Y=y| X = x, Z=z)P(z)$$

-   No notion of a linear control or anything like that; idea really is like blocking: think lots of discrete data and no missing patterns
-   Note this is a formula for a (possibly counterfactual) *level*; a counterfactual difference would be given in the obvious way by:

$$P(Y=y | \hat{x}) - P(Y=y | \hat{x}')$$

### Backdoor Proof {.smaller}

Following Pearl (2009), Chapter 11. Let $T$ denote the set of parents of $X$: $T := pa(X)$, with (possibly vector valued) realizations $t$. These might not all be observed.

If the backdoor criterion is satisfied, we have:

1.  $Y$ is independent of $T$, given $X$ and observed data, $Z$ (since $Z$ blocks backdoor paths)
2.  $X$ is independent of $Z$ given $T$. (Since $Z$ includes only nondescendents)

-  **Key idea**: The intervention level relates to the observational level as follows: $$p(y|\hat{x}) = \sum_{t\in T} p(t)p(y|x, t)$$

-  Think of this as *fully accounting for the (possibly unobserved) causes of $X$, $T$*

### Backdoor Proof {.smaller}

We want to get to:

$$p(y|\hat{x}) = \sum_{t\in T} p(t)p(y|x, t)$$

-   But of course we do not observe $T$, rather we observe $Z$. So we now need to write everything in terms of $Z$ rather than $T$.

We bring $Z$ into the picture by writing: 

$$p(y|\hat{x}) = \sum_{t\in T} p(t) \sum_z p(y|x, t, z)p(z|x, t)$$


now we want to get rid of $T$...

### Backdoor Proof {.smaller}

now we want to get rid of $T$...

-   Using the two conditions from the backdoor definition above:

    1.  replace $p(y|x, t, z)$ with $p(y | x, z)$
    2.  replace $p(z|x, t)$ with $p(z|t)$

This gives: $$p(y|\hat x) = \sum_{t \in T} p(t) \sum_z p(y|x, z)p(z|t)$$

Cleaning up, we can get rid of $T$: 

$$p(y|\hat{x}) =  \sum_z p(y|x, z)\sum_{t\in T} p(z|t)p(t) = \sum_z p(y| x, z)p(z)$$


### Backdoor proof figure {.smaller}

For intuition: 

```{r, echo = FALSE, fig.height = 4, fig.width = 10}

make_model(" Y <- Z <- T -> X -> Y ") |> plot(x_coord = c(1, 2, 2, 3), y_coord = c(1, 1, 2, 1))

```

We would be happy if we could condition on the parent $T$, but $T$ is not observed. However we can use $Z$ instead making use of the fact that:

1.  $p(y|x, t, z) = p(y | x, z)$ (since $Z$ blocks)
2.  $p(z|x, t) = p(z|t)$ (since $Z$ is upstream and blocked by parents, $T$)

### Adjustment criterion {.smaller}

See @shpitser2012validity

The adjustment criterion is satisfied by $Z$ (relative to $X$, $Y$) if:

1.  no element of $Z$ is a descendant in the mutilated graph of any variable $W\not\in X$ which lies on a proper causal path from $X$ to $Y$
2.  $Z$ blocks all **noncausal paths** from $X$ to $Y$

Note:

-   mutilated graph: remove arrows pointing into $X$
-   proper pathway: A *proper* causal pathway from $X$ to $Y$ only intersects $X$ at the endpoint

### These are different. Simple illustration.

Here $Z$ satisfies the adjustment criterion but not the backdoor criterion:

```{r, echo = FALSE, fig.width = 8, fig.height = 3,  fig.align="center", out.width='.9\\textwidth'}
par(mar=c(1,1,1,1))
hj_ggdag(x = c(0,1,1),
       y = c(1,0,2),
       names = c( "X", "Z",  "Y" ),
       arcs = cbind( c(1, 1), c(2, 3)),
       title = "Controlling for Z is OK",
       labelparse = FALSE,
       padding = .4, contraction = .15) 

```

$Z$ is descendant of $X$ but it is not a descendant of a node on a path from $X$ to $Y$. No harm adjusting for $Z$ here, but not necessary either.

<!-- No solution here: -->

<!-- $$U \rightarrow X \rightarrow Z \rightarrow Y, U \rightarrow Y$$ -->

<!-- Note here there is a non-causal but also not backdoor path from $X$ to $Z$ ($Z\rightarrow Y \leftarrow U \rightarrow X$). As @shpitser2012validity note "while all back-door paths are non-causal, not all -->

<!-- non-causal paths are back-door" -->

### Frontdoor criterion (Pearl) {.smaller}

Consider this DAG:

```{r, echo = FALSE, fig.height = 4, fig.width = 10}
make_model( "Y <-U -> X-> M -> Y") |> plot(x_coord = c(2, 1, 2, 3), y_coord = c(2,1,1,1))
```

* The relationship between $X$ and $Y$ is confounded by $U$. 
* However the $X\rightarrow Y$ effect is the product of the $X\rightarrow M$ effect and the $M\rightarrow Y$ effect

Why?

### Identification through the front door {.smaller}


If:

* $M$ (possibly a set) blocks all *directed* paths from $X$ to $Y$
* there is no backdoor path $X$ to $M$
* $X$ blocks all backdoor paths from $M$ to $Y$
and
* all ($m,z$) combinations arise with positive probability

Then $\Pr(y| \hat x)$ is identifiable and given by:

$$\Pr(y| \hat x)  = \sum_m\Pr(m|x)\sum_{x'}\left(\Pr(y|m,x')\Pr(x')\right)$$

### Frontdoor criterion (Proof) {.smaller}

We want to get $\Pr(y | \hat x)$

From the graph the joint distribution of variables is:

$$\Pr(x,m,y,u)  = \Pr(u)\Pr(x|u)\Pr(m|x)\Pr(y|m,u)$$
If we intervened on $X$ we would have ($\Pr(X = x |u)=1$):

$$\Pr(m,y,u | \hat x)  = \Pr(u)\Pr(m|x)\Pr(y|m,u)$$
If we sum up over $u$ and $m$ we get:

$$\Pr(m,y| \hat x)  = \Pr(m|x)\sum_u\left(\Pr(y|m,u)\Pr(u)\right)$$
$$\Pr(y| \hat x)  = \sum_m\Pr(m|x)\sum_u\left(\Pr(y|m,u)\Pr(u)\right)$$

The first part is fine; the second part however involves $u$ which is unobserved. So we need to get the $u$ out of $\sum_u\left(\Pr(y|m,u)\Pr(u)\right)$.

### Frontdoor criterion {.smaller}

Now, from the graph:

1. $M$ is d-separated from $U$ by $X$:

$$\Pr(u|m, x) = \Pr(u|x)$$
2. $X$ is d-separated from $Y$ by $M$, $U$

$$\Pr(y|x, m, u) = \Pr(y|m,u)$$
That's enough to get $u$ out of $\sum_u\left(\Pr(y|m,u)\Pr(u)\right)$

### Frontdoor criterion {.smaller}


$$\sum_u\left(\Pr(y|m,u)\Pr(u)\right) = \sum_x\sum_u\left(\Pr(y|m,u)\Pr(u|x)\Pr(x)\right)$$

Using the 2 equalities we got from the graph:

$$\sum_u\left(\Pr(y|m,u)\Pr(u)\right) = \sum_x\sum_u\left(\Pr(y|x,m,u)\Pr(u|x,m)\Pr(x)\right)$$

So:

$$\sum_u\left(\Pr(y|m,u)\Pr(u)\right) = \sum_x\left(\Pr(y|m,x)\Pr(x)\right)$$

Intuitively: $X$ blocks the back door between $Z$ and $Y$ just as well as $U$ does

### Frontdoor criterion {.smaller}

Substituting we are left with:

$$\Pr(y| \hat x)  = \sum_m\Pr(m|x)\sum_{x'}\left(\Pr(y|m,x')\Pr(x')\right)$$

(The $'$ is to distinguish the $x$ in the summation from the value of $x$ of interest)

It's interesting that $x$ remains in the right hand side in the calculation of the $m \rightarrow y$ effect, but this is because $x$ blocks a backdoor from $m$ to $y$

### Front foor {.smaller}

Bringing all this together into a claim we have:


If:

* $M$ (possibly a set) blocks all *directed* paths from $X$ to $Y$
* there is no backdoor path $X$ to $M$
* $X$ blocks all backdoor paths from $M$ to $Y$
and
* all ($m,z$) combinations arise with positive probability

Then $\Pr(y| \hat x)$ is identifiable and given by:

$$\Pr(y| \hat x)  = \sum_m\Pr(m|x)\sum_{x'}\left(\Pr(y|m,x')\Pr(x')\right)$$

### Front foor

* This is a very elegant and surprising result
* There are not many obvious applications of it however
* The conditions would be violated for example if unobserved third things caused both $M$ and $Y$

## In code: Dagitty

There is a package [@dagitty] for figuring out what to condition on.

```{r}
library(dagitty)
```

### In code: Dagitty

Define a dag using dagitty syntax:

```{r}
g <- dagitty("dag{X -> M -> Y ; Z -> X ; Z -> R -> Y}")
```

There is then a simple command to check whether two sets are d-separated by a third set:

```{r, comment = NA}
dseparated(g, "X", "Y", "M")
dseparated(g, "X", "Y", c("Z","M"))
```

### Dagitty: Find adjustment sets

And a simple command to identify the adjustments needed to identify the effect of one variable on another:

```{r, comment = NA}
adjustmentSets(g, exposure = "X", outcome = "Y")
```

### Important Examples : Confounding

Example where $Z$ is correlated with $X$ and $Y$ and is a confounder

```{r, echo = FALSE, fig.width = 8, fig.height = 6,  fig.align="center", out.width='.9\\textwidth'}
par(mar=c(1,1,1,1))
hj_ggdag(x = c(-1, 0, -1, 1, 3),
       y = c(0, 2, 4, 1, 2),
       names = c( " ", "Z", " ", "X", "Y" ), 
       arcs = cbind( c(2, 2, 4),
                     c(4, 5, 5)),
       title = "Controlling for Z can remove bias",
       labelparse = FALSE,
       padding = .4, contraction = .15) 

```

### Confounding

Example where $Z$ is correlated with $X$ and $Y$ but it is *not* a confounder

```{r, echo = FALSE, fig.width = 8, fig.height = 6,  fig.align="center", out.width='.9\\textwidth'}
par(mar=c(1,1,1,1))
hj_ggdag(x = c(-1, 0, -1, 1, 3),
       y = c(0, 2, 4, 1, 2),
       names = c("U2",
         "Z", "U1", "X", "Y" ),
       arcs = cbind( c(1, 3, 1, 3, 4),
                     c(2, 2, 4, 5, 5)),
       title = "Unbiased without controlling for Z",
       padding = .4, contraction = .15,
       labelparse = FALSE) 

```

### Important Examples : Collider

But controlling can also cause problems. In fact conditioning on a temporally pre-treatment variable could cause problems. Who'd have thunk? Here is an example from Pearl (2005):

```{r, echo = FALSE, fig.width = 8, fig.height = 6,  fig.align="center", out.width='.9\\textwidth'}
par(mar=c(1,1,1,1))
hj_ggdag(x = c(-1, 0, -1, 1, 3),
       y = c(0, 2, 4, 1, 2),
       names = c("U2", "Z", "U1", "X", "Y" ),
       arcs = cbind( c(1, 3, 1, 3, 4),
                     c(2, 2, 4, 5, 5)),
       title = "Controlling for Z can induce bias",
       padding = .4, contraction = .15,
       labelparse = FALSE) 

```

### Illustration of identification failure from conditioning on a collider {.smaller}

::: smaller
```{r}
#| results: asis

U1 <- rnorm(10000);  U2 <- rnorm(10000)
Z  <- U1+U2
X  <- U2 + rnorm(10000)/2
Y  <- U1*2 + X

lm_robust(Y ~ X) |> tidy() |> kable(digits = 2)
lm_robust(Y ~ X + Z) |> tidy() |> kable(digits = 2)
```
:::

### Let's look at that in dagitty

```{r, comment = NA}
g <- dagitty("dag{U1 -> Z  ; U1 -> y ; U2 -> Z ; U2 -> x  -> y}")
adjustmentSets(g, exposure = "x", outcome = "y")
isAdjustmentSet(g, "Z", exposure = "x", outcome = "y")
isAdjustmentSet(g, NULL, exposure = "x", outcome = "y")
```

Which means, no need to condition on anything.

<!-- g2 <- backDoorGraph(g) -->

<!-- dseparated(g2, "x", "y") -->

<!-- dseparated(g2, "x", "y", "Z") -->

### Collider & Confounder

A bind: from Pearl 1995.

```{r, echo = FALSE, fig.width = 8, fig.height = 6,  fig.align="center", out.width='.9\\textwidth'}
par(mar=c(1,1,1,1))
hj_ggdag(x = c(-1, 0, -1, 1, 3),
       y = c(0, 2, 4, 1, 2),
       names = c("U2",
         "Z", "U1", "X", "Y" ),
       arcs = cbind( c(1, 3, 1, 3, 4, 2),
                     c(2, 2, 4, 5, 5, 4)),
       title = "Z is a confound but controlling for it can induce bias",
       padding = .4, contraction = .15, labelparse = FALSE) 

```

For a solution for a class of related problems see @robins2000marginal

### Let's look at that in dagitty {.smaller}

```{r, comment = NA}
g <- dagitty("dag{U1 -> Z  ; U1 -> y ; 
             U2 -> Z ; U2 -> x  -> y; 
             Z -> x}")
adjustmentSets(g, exposure = "x", outcome = "y")
```

which means you have to adjust on an unobservable. Here we double check that including or not including "Z" is enough:

```{r, comment = NA}
isAdjustmentSet(g, "Z", exposure = "x", outcome = "y")
isAdjustmentSet(g, NULL, exposure = "x", outcome = "y")
```

### Collider & Confounder

So we cannot *identify* the effect here. But can we still learn about it?
