---
format: 
   revealjs:
      embed-resources: true
      theme: serif
      slide-level: 3
      slide-number: true
      toc-depth: 2
      show-slide-number: all
      preview-links: auto
      number-sections: true
      link-color: orange
title: "Lectures on causal inference and experimental methods"
author: "Macartan Humphreys"
bibliography: bib.bib
---


# Roadmap

1. Intro | [Course outline, tools](#secoutline) |  [`DeclareDesign`](#secdd)
2. Causality | [Fundamental problems and solutions](#seccausality) | [Inquiries and identification](#secestimands) 
3. Estimation and Inference 1:  [Frequentist](#secfisher) 
4. Estimation and Inference 2: [Bayesian](#secbayes)
5. Design |  [Experimental Design](#secdesign) |  [Design evaluation](#secdiagnosis)
6. Topics 1| [Topics and techniques 1](#citopics1) 
7. Topics 2| [Topics and techniques 2](#citopics2) 



```{css}
##| echo: false
##| include: false

.reveal table {
  font-size: medium;
}

## also: smaller, small

```
  
```{r, include = FALSE}

source("setup.R")
options(future.globals.maxSize = 3 * 1024^3)  # 2GB

# remotes::install_github("jhelvy/renderthis")
# renderthis::to_pdf("ci_2024.html")
```

```{r, child = "1.1_intro.qmd"}

```


# `DeclareDesign` {#secdd}

*How to define and assess research designs* 

```{r, child = "1.2_declaredesign.qmd"}

```


```{r, child = "2.1_causality.qmd"}

```

# Inquiries {#secestimands}

*Well posed questions*

```{r, child = "2.2_estimands.qmd"}

```

# Frequentist Analysis {#secfisher}

*Estimation and testing*

```{r, child = "3_fisher.qmd"}

```


```{r, child = "4_bayes.qmd"}

```


```{r, child = "5.1_design.qmd"}

```


```{r, child = "5.2_evaluation.qmd"}

```


```{r, child = "6_topics_1.qmd"}

```


```{r, child = "7_topics_2.qmd"}

```


```{r, child = "8_topics_3.qmd"}

```


# References

### References
