---
format: 
   revealjs:
      embed-resources: true
      theme: serif
      slide-level: 3
      slide-number: true
      toc-depth: 2
      show-slide-number: all
      preview-links: auto
      number-sections: true
      link-color: orange
title: "Lectures on the design and analysis of experiments"
author: "Macartan Humphreys"
bibliography: bib.bib
---


# Roadmap

1. Intro: | [Course outline, tools](#intro)
1. Experimenting: | [questions, workflow, ethics](#experimenting)
2. Causality: [Fundamental problems, questions, and answers](#causality)  
3. Estimation:  [Answer Strategies](#answer_strategies) 
4. Design:  [Experimental Design and Evaluation](#design) 
6. Topics:  [Specific experimental design](#experimental_topics) 



```{css}
##| echo: false
##| include: false

.reveal table {
  font-size: medium;
}

## also: smaller, small

```
  
```{r, include = FALSE}

source("setup.R")
options(future.globals.maxSize = 3 * 1024^3)  # 2GB

# remotes::install_github("jhelvy/renderthis")
# renderthis::to_pdf("ci_2024.html")
```


# Intro {#intro}

```{r, child = "1_intro_experiments.qmd"}

```


# Experimenting {#experimenting}

```{r, child = "6_experiments_topics.qmd"}

```

# Causality {#causality}


```{r, child = "2.1_causality.qmd"}

```


```{r, child = "2.2_estimands.qmd"}

```


# Causality {#answer_strategies}

```{r, child = "3_fisher.qmd"}

```


# Design {#design}

```{r, child = "1.2_declaredesign.qmd"}

```


```{r, child = "5.1_design.qmd"}

```


```{r, child = "5.2_evaluation.qmd"}

```

# Topics {#topics}

```{r, child = "6_experiments_topics.qmd"}

```




# References

### References
