---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: false
title: "Group 4 Exercises"
author:
  - Nader
  - Philipp
  - Yoanna
  - Robert
  - Daniel
---

```{r}
# packages
pacman::p_load(tidyverse,
               declaredesign)
```


# 2.4

Declare a design in which:

- The assignment of a treatment **X** depends in part on another binary variable **W**. Specifically:
  - Pr(X = 1 | W = 0) = 0.2
  - Pr(X = 1 | W = 1) = 0.5
- The outcome **Y** depends on both **X** and **W**. Specifically, Y = X * W + u, where **u** is a random shock.

---

- Diagnose the design with three approaches to estimating the effect of **X** on **Y**:
  1. **Ignoring W**: Estimate the effect of **X** on **Y** without considering **W**.
  2. **Adding W as a Linear Control**: Include **W** as a linear control in the estimation.
  3. **Interaction between W and X**: Include both **W** and an interaction term between **W** and **X** in the estimation.

Discuss the results. Do any of these approaches return the correct answer?

## Option 1: Ignoring W

```{r option_1, echo = TRUE}
# design here
```

## Option 2: Adding W as a Linear Control

```{r option_2, echo = TRUE}
# design here
```

## Option 3: Interaction between W and X

```{r option_3, echo = TRUE}
# design here
```

# 3.4

# 4.4



