---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
    code-fold: show
    code-summary: "Show the code"
    scrollable: true
    echo: true
title: "Exercise 4.2"
author: "Eduard, Felix, Linus & Noah"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---


```{r setup, echo=FALSE}
# install and load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  DeclareDesign,
  dagitty,
  tidyverse,
  here,
  kableExtra
  )

# declare file location
here::i_am("4.2.qmd")
```


# Puzzle

* Draw a DAG with 5 nodes representing  a situation in which $X$ causes $Y$ though $M$, $C$ affects both $X$ and $M$ and $D$ affects both $M$ and $Y$.
* Think through what  set of nodes which, when controlled for, would allow for the identification of the effect of $X$ on $Y$. 
* Represent it in `dagitty` and check your answer
* Bonus: Declare the design and compare the behavior of designs that do and do not control for these nodes.

# Solution

## Define causal paths between variables

```{r}
dag <- dagitty("dag { X -> M -> Y C -> X C -> M D -> M }")
```


## DAG

```{r}
plot(dag)
```


## Find adjustment set

```{r}
adjustment_set <- adjustmentSets(dag, exposure = "X", outcome = "Y")
adjustment_set
```


## DeclareDesign

```{r}
design_4_2 <- 
  declare_model(
    N = 10000,
    U = rnorm(N),
    C = rnorm(N),
    D = rnorm(N),
    M = rnorm(N) + 0.5 * C + 0.2 + D,
    X = rbinom(N, 1, prob = pnorm(0.8 * C)),
    Y_X_0 = X * M + D + U,
    Y_X_1 = Y_X_0 + 0.5
  )  +
  declare_inquiry(ATE = mean(Y_X_1 - Y_X_0)) +
  declare_measurement(Y = reveal_outcomes(Y ~ X)) +
  declare_estimator(
    Y ~ X,
    label = "naive"
  ) +
  declare_estimator(
    Y ~ X + C,
    label = "right control"
  ) +
  declare_estimator(
    Y ~ X + D,
    label = "wrong control"
  )
```


## Look at the design

```{r}
#| code-fold: true
design_4_2 |> 
  draw_data() |> 
  select(ID, U, C, D, M, X, Y) |> 
  pivot_longer(
    cols = c(U, C, D, M, X, Y),
    names_to = "variable",
    values_to = "value"
  ) |> 
  ggplot() +
  geom_histogram(aes(value)) +
  facet_wrap(~variable)
```

## Run design

```{r}
design_4_2 |> 
  run_design() |> 
  kable(digits = 2) |> 
  kable_styling(font_size = 20)
```

## Diagnose design

```{r}
design_4_2 |> 
  diagnose_designs() |> 
  reshape_diagnosis() |> 
  kable(digits = 2) |> 
  kable_styling(font_size = 20)
```