---
title: "In class exercises: RI and assignment schemes"
format: pdf
---

This exercise connects intuitions for randomization inference with the implications of different assignmet schemes?  

```{r setup, include = FALSE}
library(DeclareDesign)
library(knitr)
```


You have the following data:

```{r, echo = FALSE}
fabricate(N = 8,
  B = rep(0:1, each = 4),
  C = rep(1:4, each = 2),
  Z = c(0,0,1,1,0,0,1,1),
  Y = Z) |> kable()
```

Calculate the $p$ value for the sharp null hypothesis that effects of $Z$ on $Y$ are zero for all units under each of these conditions:

1. $Z$ was assigned using **complete randomization**. 
2. $Z$ was assigned using **block randomization**, using $B$ as a block.
3. $Z$ was assigned using **cluster randomization**, where $C$ denotes clusters.
4. $Z$ was assigned using **block cluster randomization**, using $B$ for blocks and $C$ for clusters.

Implement one sided tests in all cases.

 
*Discuss*: we usually expect that blocking will improve power but clustering will reduce it. Is the logic in tension with what you just found. Explain why or why not.

