---
title: "Untitled"
format: html
editor: visual
---

**9.1 Spillovers**

Imagine a study with three subjects. Each subject’s potential outcomes are as follows:

-   0 if in control

-   n if in treatment when n subjects are assigned to treatment

So for example if one unit is in treatment that unit has outcome Y=1, and the others have Y = 0; if all 3 are in treatment they all have Y=3.

1.  Write down the potential outcomes for all possible assignments, including all and none assigned o treatment

2.  Write down the difference in means calculation for each possible assignment

3.  Define two estimands of interest.

4.  Define a randomization scheme and answer strategy that will return an unbiased estimate of each estimand.

| Treatment Assignments | Subject 1 | Subject 2 | Subject 3 | Diff in Means   |
|-----------------------|-----------|-----------|-----------|-----------------|
| **Y_0_0_0**           | 0         | 0         | 0         | (Control Group) |
| **Y_0_0_1**           | 0         | 0         | 1         | 1/3             |
| **Y_0_1_0**           | 0         | 1         | 0         | 1/3             |
| **Y_0_1_1**           | 0         | 2         | 2         | 4/3             |
| **Y_1_0_0**           | 1         | 0         | 0         | 1/3             |
| **Y_1_0_1**           | 2         | 0         | 2         | 4/3             |
| **Y_1_1_0**           | 2         | 2         | 0         | 4/3             |
| **Y_1_1_1**           | 3         | 3         | 3         | 3               |

One of these is the direct average treatment effect:

\[Y_1_0_0 + Y_1_0_1 + Y_1_1_0 + Y_1_1_1\]/4 - \[Y_0_0_0 + Y_0_0_1 + Y_0_1_0 + Y_0_1_1\]/4

= 10/4 - 3/4 = 7/4 = 1.75

\[Y_1_0_0 + Y_1_0_1 + Y_1_1_0 + Y_1_1_1\]/4 - \[Y_0_0_0\]

= 10/4 - 0 = 2.5

ATT might also be an estimand of interest...?

1.  Define a randomization scheme and answer strategy that will return an unbiased estimate of each estimand.

Not sure.

#3.  Define two estimands of interest.



```{r}
library(DeclareDesign)
model<- declare_model(N<-3,
                      U = rnorm(N),
                      n<- sample(Z=1),
                      spillover<- sample(n=3)
                      potential_outcomes(Y~b*spillover +runif(1, 0, 0.5)*Z +rnorm(N)))
# 1. interest of estimands: ATE
inquiry_ate <- declare_inquiry(ATE = mean(Y_Z_1)-mean(Y_Z_0))
  
  
# 2. insterest of estimands:
inquiry_att <- declare_inquiry(ATT = mean(Y_Z_1)-mean(Y_Z_0), if Z=1)
    
  
```
#4.  Define a randomization scheme and answer strategy that will return an unbiased estimate of each estimand.

```{r}
data_strategy <- 
  declare_sampling(S=complete_rs(N=N, n=3), filter =S ==1)+
  declare_assignment(Z=complete_ra(N))+
  declare_measurement(Y=reveal_outcomes(Y~Z))


answer_strategy_ate <- declare_estimator(Y~Z, .method = difference_in_means, inquiry="ATE")
answer_strategy_att <- declare_estimator(Y~Z, .method = difference_in_means, inquiry="ATT")

declaration <- model + inquiry_ate + inquiry_att + data_strategy + answer_strategy_ate+ answer_strategy_att
diagnose_design(declaration, diagnosands = diagnosands)
```
