---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
title: "Exercise 4.1 & 5.1"
author: "Nora Chirikure, Ziyu Lin, Felix Hüskes, Anton Bogs & Elias Koch"
---
```{r setup, include=FALSE}
pacman::p_load(knitr, tidyverse, CausalQueries, DeclareDesign, rvest, kableExtra, knitr, ggdag) #load packages

table_5_1 <- html_nodes(read_html("https://macartan.github.io/ci/exercises.html#estimation-and-inference-frequentist"), #pull data for 5.1
                    "table") 
```


# Collider Bias (4.1)
Declare a simple design in which 

- X and Y both have a positive effect on (binary) K but X does not cause Y 
- a researcher conditions on K == 1 when estimating the effect of X on Y

Show that this can generate biased results. ***Can you find situations where the bias can be either positive or negative?***

## DAG
```{r}
set.seed(12) 
dagify(K ~ X,
       K ~ Y,
       outcome = "K",
       exposure = c("X", "Y"),
       labels = c("K" = "K==1")) |> 
  ggdag(use_labels = "label")+
  theme_dag()
```

## How can the relationship between X and Y be changed?

###  *K = rbinom(N, 1, prob = pnorm(X + Y))*
```{r}
M1 <- 
  declare_model(
    N = 100000, 
    X = rnorm(N),
    Y = rnorm(N),
    K = rbinom(N, 1, prob = pnorm(X + Y))) +
  declare_inquiry(Ybar = mean(Y[X==1])) +
  declare_sampling(S = simple_rs(N, prob = 0.2), 
                   filter = S == 1) +
  declare_estimator(
    Y ~ X,
    .method = lm_robust,
    .summary = tidy,
    term = "X",
    inquiry = "ATE",
    label = "OLS",
    subset = K == 1
  )

draw_data(M1)|> ggplot(aes(X, Y, color =K)) + geom_point()
```

###  *K = rbinom(N, 1, prob = pnorm(X x Y))*
```{r}
M2 <- 
  declare_model(
    N = 100000, 
    X = rnorm(N),
    Y = rnorm(N),
    K = rbinom(N, 1, prob = pnorm(X * Y))) +
  declare_inquiry(Ybar = mean(Y[X==1])) +
  declare_sampling(S = simple_rs(N, prob = 0.2), 
                   filter = S == 1) +
  declare_estimator(
    Y ~ X,
    .method = lm_robust,
    .summary = tidy,
    term = "X",
    inquiry = "ATE",
    label = "OLS",
    subset = K == 1
  )

draw_data(M2)|> ggplot(aes(X, Y, color =K)) + geom_point()

```


### *K = rbinom(N, 1, prob = pnorm(X / Y))*
```{r}
M3 <- 
  declare_model(
    N = 100000, 
    X = rnorm(N),
    Y = rnorm(N),
    K = rbinom(N, 1, prob = pnorm(X / Y))) +
  declare_inquiry(Ybar = mean(Y[X==1])) +
  declare_sampling(S = simple_rs(N, prob = 0.2), 
                   filter = S == 1) +
  declare_estimator(
    Y ~ X,
    .method = lm_robust,
    .summary = tidy,
    term = "X",
    inquiry = "ATE",
    label = "OLS",
    subset = K == 1
  )

draw_data(M3)|> ggplot(aes(X, Y, color =K)) + geom_point()
```

# Estimators (5.1)
```{r}
table_data <- html_table(table_5_1, fill = TRUE)[[5]] %>%
  rename("Treat" = "Z",
         "Outcome" = "Y")
kable(table_data, "html") %>%
  kable_styling() %>%
  scroll_box(width = "100%")
```

## Block-Wise Estimated Effects
```{r}
# Calculating ATE (ATT and ATU would require pre-treatment observations)
mean_treatment_1 <- mean(table_data$Outcome[table_data$Block == "1" & 
                                              table_data$Treat == "1"])
mean_control_1 <- mean(table_data$Outcome[table_data$Block == "1" & 
                                            table_data$Treat == "0"])

mean_treatment_2 <- mean(table_data$Outcome[table_data$Block == "2" & 
                                              table_data$Treat == "1"])
mean_control_2 <- mean(table_data$Outcome[table_data$Block == "2" & 
                                            table_data$Treat == "0"])

te_1 <- mean_treatment_1 - mean_control_1
te_2 <- mean_treatment_2 - mean_control_2



# Plot Treatment Effects
df_te <- data.frame(Block = c("Block 1", "Block 2"),
                     te = c(te_1, te_2))

plot_te <- ggplot(df_te, aes(x = Block, y = te)) +
  geom_point(aes(size = 3)) +
  scale_size(guide = "none") +
  lims(y = c(0, 1.2)) +
  labs(x = "",
       y = "Estimated Effect of Z")
plot_te

```

## ATE, ATT, ATU, DIM
```{r}
ate <- #ATE
  (3/9)* #Share of observations in Block 1
  te_1+
  (6/9)* # Share of observations in Block 2
  te_2

atc <- #ATC
  (2/5)* #Share of control observations in Block 1
  te_1+
  (3/5)* #Share of control observations in Block 2
  te_2

att <- # ATT
  (1/4)* #Share of treated observations in Block 1
  te_1+
  (3/4)* #Share of treated observations in Block 2
  te_2

dim <- #DIM
  (3/4)-
  (1/5)


measures <- data.frame(Measures = c("Average Treatment\nEffect (ATT)", "Average Treatment Effect\non the Control (ATC)", "Average Treatment Effect\non the Treated (ATT)", "Difference in\nMeans (DIM)"),
                       Effect = c(ate, atc, att, dim))

ggplot(measures, aes(x = Measures %>% reorder(Effect), y = Effect))+
  geom_col()+
  labs(x = "")
```

## Redesinging & Diagnosing the last Design

```{r}
df <- data.frame(Block = c(1,1,1,2,2,2,2,2,2),
                   Z = c(0,0,1,0,0,0,1,1,1),
                   Y = c(0,0,1,0,0,1,0,1,1)) |>
  group_by(Block) |> 
  mutate(p = mean(Z), ipw = Z/p + (1-Z)/(1-p)) |>
  ungroup() |> data.frame()

design <- declare_model(data = df) +
  declare_inquiry(ATE = 5/9) +
  declare_estimator(Y ~ Z + Block, label = "FE") +
  declare_estimator(Y ~ Z, weights = ipw, label = "IPW") +
  declare_estimator(Y ~ Z, covariates = ~ Block, .method = lm_lin, label = "lin") 

diagnosands <- declare_diagnosands(
                  bias = mean(estimate - estimand),
                  sd_estimate = sd(estimate),
                  mean_se = mean(std.error))

diagnose_design(design, diagnosands = diagnosands, sims = 2)

replicate <- function(df, w) df |> uncount(w) # uncount used here to replicate data

design |> 
  insert_step(after = 1, declare_measurement(handler = replicate, w = k)) |>
  redesign(k = 1:5) |> 
  diagnose_design(diagnosands = diagnosands, sims = 2) |> tidy() |>
  ggplot(aes(k, estimate, color = estimator)) + geom_line() + facet_grid(~diagnosand, scales = "free")

```
