---
title: "Topics"
subtitle: "LATE, encouragements, compliance, and downstream experiments"
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
    link-color: orange
author: "Macartan Humphreys"
bibliography: ../bib.bib
---

```{r, include = FALSE}
source(here::here("slides", "setup.R"))
```


## Noncompliance and the LATE estimand {#LATE}

```{r, echo = FALSE}
model <- make_model("Z -> X -> Y; X <-> Y") 
model |> plot()
```

### Local Average Treatment Effects {.smaller}

Sometimes you give a medicine but only a nonrandom sample of people actually try to use it. Can you still estimate the medicine's effect?

|     | X=0                           | X=1                           |
|-----|-------------------------------|-------------------------------|
| Z=0 | $\overline{y}_{00}$ ($n_{00}$) | $\overline{y}_{01}$ ($n_{01}$) |
| Z=1 | $\overline{y}_{10}$ ($n_{10}$) | $\overline{y}_{11}$ ($n_{11}$) |

Say that people are one of 3 types:

1.  $n_a$ "always takers" have $X=1$ no matter what and have average outcome $\overline{y}_a$
2.  $n_n$ "never takers" have $X=0$ no matter what with outcome $\overline{y}_n$
3.  $n_c$ "compliers have" $X=Z$ and average outcomes $\overline{y}^1_c$ if treated and $\overline{y}^0_c$ if not.

### Local Average Treatment Effects  {.smaller}

Sometimes you give a medicine but only a nonrandom sample of people actually try to use it. Can you still estimate the medicine's effect?

|     | X=0                           | X=1                           |
|-----|-------------------------------|-------------------------------|
| Z=0 | $\overline{y}_{00}$ ($n_{00}$) | $\overline{y}_{01}$ ($n_{01}$) |
| Z=1 | $\overline{y}_{10}$ ($n_{10}$) | $\overline{y}_{11}$ ($n_{11}$) |

We can figure something about types:

|          | $X=0$                  | $X=1$                  |                                                                                                        |
|----------|:----------------------:|:----------------------:|
| $Z=0$ | $\frac{\frac{1}{2}n_c}{\frac{1}{2}n_c + \frac{1}{2}n_n} \overline{y}^0_{c}+\frac{\frac{1}{2}n_n}{\frac{1}{2}n_c + \frac{1}{2}n_n} \overline{y}_{n}$ | $\overline{y}_{a}$                                                                                                                                  |
| $Z=1$ | $\overline{y}_{n}$                                                                                                                                  | $\frac{\frac{1}{2}n_c}{\frac{1}{2}n_c + \frac{1}{2}n_a} \overline{y}^1_{c}+\frac{\frac{1}{2}n_a}{\frac{1}{2}n_c + \frac{1}{2}n_a} \overline{y}_{a}$ |

### Local Average Treatment Effects  {.smaller}

You give a medicine to 50% but only a non random sample of people actually try to use it. Can you still estimate the medicine's effect?

|         | $X=0$                                                                           | $X=1$                                                                             |
|---------|:---------------------:|:---------------------:|
| $Z=0$ | $\frac{n_c}{n_c + n_n} \overline{y}^0_{c}+\frac{n_n}{n_c + n_n} \overline{y}_n$ | $\overline{y}_{a}$ |
| \(n\) | ($\frac{1}{2}(n_c + n_n)$)                                                      | ($\frac{1}{2}n_a$)  |
| $Z=1$ | $\overline{y}_{n}$                  | $\frac{n_c}{n_c + n_a} \overline{y}^1_{c}+\frac{n_a}{n_c + n_a} \overline{y}_{a}$ |
| \(n\) | ($\frac{1}{2}n_n$)                                                              | ($\frac{1}{2}(n_a+n_c)$) |                 

Key insight: the contributions of the $a$s and $n$s  are the *same* in the $Z=0$ and $Z=1$ groups so if you difference you are left with the *changes* in the contributions of the $c$s. 


### Local Average Treatment Effects  {.smaller}

Average in $Z=0$ group: $\frac{{n_c} \overline{y}^0_{c}+ \left(n_{n}\overline{y}_{n} +{n_a} \overline{y}_a\right)}{n_a+n_c+n_n}$

Average in $Z=1$ group: $\frac{{n_c} \overline{y}^1_{c} + \left(n_{n}\overline{y}_{n} +{n_a} \overline{y}_a \right)}{n_a+n_c+n_n}$

So, the *difference* is the ITT:  $({\overline{y}^1_c-\overline{y}^0_c})\frac{n_c}{n}$

Last step:

$$ITT = ({\overline{y}^1_c-\overline{y}^0_c})\frac{n_c}{n}$$

$$\leftrightarrow$$ 

$$LATE = \frac{ITT}{\frac{n_c}{n}}= \frac{\text{Intent to treat effect}}{\text{First stage effect}}$$

### Local Average Treatment Effects (Simpler)  {.smaller}

Let $\lambda_c =$ share compliers; $\lambda_a =$ share always takers; $\lambda_n =$ share never takers.

Average $Y$ in $Z=0$ group: $$\lambda_c\overline{y}^0_{c}+ \lambda_n\overline{y}_{n} +\lambda_a\overline{y}_a$$

Average $Y$ in $Z=1$ group: $$\lambda_c\overline{y}^1_{c}+ \lambda_n\overline{y}_{n} +\lambda_a\overline{y}_a$$


So, the *difference*, or  ITT is:  

\begin{eqnarray*}
ITT &=& \lambda_c({\overline{y}^1_c-\overline{y}^0_c}) 

&=& (\text{First stage})\times LATE
\end{eqnarray*}


### The good and the bad of LATE  {.smaller}

-   You get a well-defined estimate even when there is non-random take-up
-   May sometimes be used to assess mediation or knock-on effects
-   But:
    -   You need assumptions (monotonicity and the exclusion restriction -- *where were these used above*?)
    -   Your estimate is only for a subpopulation
    -   The subpopulation is not chosen by you and is unknown
    -   Different encouragements may yield different estimates since they may encourage different subgroups


### Pearl and Chickering again

With and without an imposition of monotonicity

```{r}
#| echo: true
#| eval: false

data("lipids_data")

models <- 
  list(unrestricted =  make_model("Z -> X -> Y; X <-> Y"),
       restricted =  make_model("Z -> X -> Y; X <-> Y") |>
         set_restrictions("X[Z=1] < X[Z=0]")) |> 
  lapply(update_model,  data = lipids_data, refresh = 0) 

models |>
  query_model(query = list(CATE = "Y[X=1] - Y[X=0]", 
                           Nonmonotonic = "X[Z=1] < X[Z=0]"),
              given = list("X[Z=1] > X[Z=0]", TRUE),
              using = "posteriors") 
```


### Pearl and Chickering again

With and without an imposition of monotonicity:

```{r}
#| echo: false
#| eval: true

data("lipids_data")

if(run)
  list(unrestricted =  make_model("Z -> X -> Y; X <-> Y"),
       restricted =  make_model("Z -> X -> Y; X <-> Y") |>
         set_restrictions("X[Z=1] < X[Z=0]")) |> 
  lapply(update_model,  data = lipids_data, refresh = 0) |>
  query_model(query = list(CATE = "Y[X=1] - Y[X=0]", Nonmonotonic = "X[Z=1] < X[Z=0]"),
              given = list("X[Z=1] > X[Z=0]", TRUE),
              using = "posteriors") |>
    write_rds(here::here("slides", "saved/lipids_again.rds"))

read_rds(here::here("slides", "saved/lipids_again.rds")) |> select(model, query, mean, sd) |> arrange(query) |> kable(digits = 2)
```


In one case we *assume* monotonicity, in the other we **update** on it (easy in this case because of the empirically verifiable nature of one sided non compliance)

