---
title: "Topics"
subtitle: "RDD"
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
    link-color: orange
author: "Macartan Humphreys"
bibliography: ../bib.bib
---

```{r, include = FALSE}
source(here::here("slides", "setup.R"))
```


## Regression discontintuity {#rdd}

Errors and diagnostics



### Intuition

* The core idea in an RDD design is that if a decision rule assigns units that are *almost identical* to each other to treatment and control conditions then we can infer effects *for those cases* by looking *at those cases*.


See excellent introduction:  @lee2010regression

### Intuition  {.smaller}


* Kids born on 31 August start school a year younger than kids born on 1 September: does starting younger help or hurt?

* Kids born on 12 September 1983 are more likely to register Republican than kids born on  10 September 1983: can this identify the effects of registration on long term voting? 

* A district in which Republicans got 50.1% of the vote get a Republican representative while districts in which Republicans got 49.9% of the vote do not: does having a Republican representative make a difference for these districts?



### Argument for identification {.smaller}

Setting:

* Typically the decision is based on a value on a "running variable", $X$. e.g. Treatment if $X > 0$
* The estimand is $\mathbb{E}[Y(1) - Y(0)|X=0]$

Two arguments:

1. Continuity: $\mathbb{E}[Y(1)|X=x]$ and $\mathbb{E}[Y(0)|X=x]$ are continuous (at $x=0$) in $x$: so  $\lim_{\hat x \rightarrow 0}\mathbb{E}[Y(0)|X=\hat x] = \mathbb{E}[Y(0)|X=\hat 0]$

2. Local randomization: tiny things that determine exact values of $x$ are *as if* random and so we can think of a local experiment around $X=0$.


### Argument for identification  {.smaller}

Note: 

* continuity argument requires continuous $x$: granularity
* also builds off a conditional expectation function defined at $X=0$

Exclusion restriction is  *implicit* in continuity: If *something else* happens at the threshold then the conditional expectation functions jump at the thresholds

Implicit: $X$ is exogenous in the sense that units  cannot adjust $X$ in order to be on one or the other side of the threshold 

### Evidence

Typically researchers show:

1. "First stage" results: assignment to treatment does indeed jump at the threshold
2. "ITT": outcomes jump at the threshold
3.  LATE (if fuzzy / imperfect compliance) using IV

### Evidence

Typically researchers show:

In addition:

1. Arguments for no other treatments at the threshold
2. Arguments for no "sorting" at the threshold
3. Evidence for no "heaping" at the threshold (McCrary density test)

Sometimes: 

* argue for why estimates extend beyond the threshold
* exclude points *at* the threshold (!)

### Design

```{r}
library(rdss) # for helper functions
library(rdrobust)

cutoff <- 0.5
bandwidth <- 0.5

control <- function(X) {
  as.vector(poly(X, 4, raw = TRUE) %*% c(.7, -.8, .5, 1))}
treatment <- function(X) {
  as.vector(poly(X, 4, raw = TRUE) %*% c(0, -1.5, .5, .8)) + .25}

rdd_design <-
  declare_model(
    N = 1000,
    U = rnorm(N, 0, 0.1),
    X = runif(N, 0, 1) + U - cutoff,
    D = 1 * (X > 0),
    Y_D_0 = control(X) + U,
    Y_D_1 = treatment(X) + U
  ) +
  declare_inquiry(LATE = treatment(0) - control(0)) +
  declare_measurement(Y = reveal_outcomes(Y ~ D)) + 
  declare_sampling(S = X > -bandwidth & X < bandwidth) +
  declare_estimator(Y ~ D*X, term = "D", label = "lm") + 
  declare_estimator(
    Y, X, 
    term = "Bias-Corrected",
    .method = rdrobust_helper,
    label = "optimal"
  )

```

### RDD Data plotted

Note `rdrobust` implements:

* local polynomial Regression Discontinuity (RD) point estimators
* robust bias-corrected confidence intervals 

See  @calonico2014robust and related papers `? rdrobust::rdrobust`

### RDD Data plotted

```{r}
rdd_design  |> draw_data() |> ggplot(aes(X, Y, color = factor(D))) + 
  geom_point(alpha = .3) + theme_bw() +
  geom_smooth(aes(X, Y_D_0)) + geom_smooth(aes(X, Y_D_1)) + theme(legend.position = "none")

```

### RDD diagnosis

```{r, eval = FALSE, echo = TRUE}
rdd_design |> diagnose_design()
```


```{r, echo = FALSE, eval = TRUE}
if(run)
  rdd_design |> diagnose_design() |> write_rds(here::here("slides", "saved/rdd_diagnosis.rds"))

read_rds(here::here("slides", "saved/rdd_diagnosis.rds")) |> reshape_diagnosis() |> 
  select(Estimator, 'Mean Estimate', 'Bias', 'SD Estimate', 'Coverage') |> kable()
```

### Bandwidth tradeoff {.smaller}

```{r, echo = TRUE, eval = FALSE}
rdd_design |> 
  redesign(bandwidth = seq(from = 0.05, to = 0.5, by = 0.05)) |> 
  diagnose_designs()
```

```{r, echo = FALSE, eval = TRUE, fig.height = 2, fig.width = 8}

if(run)
rdd_design |> 
  redesign(bandwidth = seq(from = 0.1, to = 0.5, by = 0.05)) |> 
  diagnose_designs() |>
  write_rds(here::here("slides", "saved/rddbandwidth.rds"))

read_rds(here::here("slides", "saved/rddbandwidth.rds"))|>  tidy() |>
  filter(diagnosand %in% c("bias", "rmse", "sd_estimate")) |>
  ggplot(aes(bandwidth, estimate, color = diagnosand)) + geom_line()  + theme_bw() +
  facet_grid(~estimator)

```

* As we increase the bandwidth, the lm bias gets worse, but slowly, while the error falls. 
* The best bandwidth is relatively wide. 
* This is more true for the optimal estimator.


### Geographic RDs {.smaller}

Are popular in political science:

* Put a lot of pressure on assumption of no alternative treatment---including "random" country level shocks!
* Put a lot of pressure on no sorting assumptions (why was the border put where it was; why did units settle here or there?)
* Put a lot of pressure on SUTVA: people on one side are literally proximate to people on another


See @keele2015geographic




