---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
title: "Topics"
subtitle: "survey experiments"
author: "Macartan Humphreys"
bibliography: ../bib.bib
---

  
```{r, include = FALSE}

source("setup.R")

```


## Survey experiments {#survey-experiments}


* Survey experiments are used to *measure* things: nothing (except answers) should be *changed*!
* If the experiment in the survey is changing things then it is a field experiment in a survey, not a survey experiment


### The list experiment: Motivation

* Multiple survey experimental designs have been generated to make it easier for subjects to answer **sensitive questions**

* The key idea is to use inference rather than measurement.

* Subjects are placed in different conditions and the conditions *affect* the **answers** that are given in such a way that you can infer some underlying quantity of interest

### The list experiment: Motivation

```{r, echo = FALSE, caption = "Questions and values collide in an answer"}
make_model("Z -> Question -> Answer <- Value") |> plot(textcol = "black", nodecol = "lightgrey")

```

This is an obvious DAG but the main point is to be clear that the *Value* is the quantity of interest and the value is not affected by the treatment, *Z*.

### The list experiment: Motivation

The list experiment supposes that:

1. Subjects do not want to give a direct answer to a question
2. They nevertheless are willing to truthfully answer an indirect question

**In other words**: sensitivities notwithstanding, *they are happy for the researcher to make correct inferences about them or their group*


### The list experiment: Strategy {.smaller}

* Respondents are given a short list and a long list.

* The long list differs from the short list in having *one extra item*---the sensitive item

* We ask how many items in each list does a respondent agree with: 

  * $Y_i(0)$ is the number of elements on a short list that a respondent agrees with
  * $Y_i(1)$ is the number of elements on a long list that a respondent agrees with
  * $Y_i(1) - Y_i(0)$ is an indicator for whether an individual agrees with the sensitive item
  * $\mathbb{E}[Y_i(1) - Y_i(0)]$ is the share of people agreeing with sensitive item



### The list experiment: Simplified example {.smaller}

How many of these do you agree with:

|        | Short list       | Long list          | "Effect"  |
|------|------------------|----------------|--------|
|        | "2 + 2 = 4"    | "2 + 2 = 4"                |   |
|        | "2 * 3 = 6"    | "2 * 3 = 6"  |   |
|        | "3 + 6 = 8"    | "Climate change is real" |   |
|        |            | "3 + 6 = 8"                |   |
| Answer | Y(0) = 2   | Y(1) = 4    |  Y(1) - Y(0) = 2 |


[Note: this is obviously not a good list. Why not?]

### The list experiment: Design

```{r}

declaration_17.3 <-
  declare_model(
    N = 500,
    control_count = rbinom(N, size = 3, prob = 0.5),
    Y_star = rbinom(N, size = 1, prob = 0.3),
    potential_outcomes(Y_list ~ Y_star * Z + control_count) 
  ) +
  declare_inquiry(prevalence_rate = mean(Y_star)) +
  declare_assignment(Z = complete_ra(N)) + 
  declare_measurement(Y_list = reveal_outcomes(Y_list ~ Z)) +
  declare_estimator(Y_list ~ Z, .method = difference_in_means, 
                    inquiry = "prevalence_rate")

diagnosands <- declare_diagnosands(
  bias = mean(estimate - estimand),
  mean_CI_width = mean(conf.high - conf.low)
)
```


### Diagnosis

```{r, eval = FALSE}
diagnose_design(declaration_17.3, diagnosands = diagnosands)
```

```{r, echo = FALSE}
if(run)
  diagnose_design(declaration_17.3, diagnosands = diagnosands) |>
  write_rds("saved/list1.rds")

read_rds("saved/list1.rds") |>
  reshape_diagnosis() |> select("Design", "Inquiry", "Bias", "Mean CI Width") |> kable()
  

```

### Tradeoffs: is the question really sensitive?


```{r}
declaration_17.4 <- 
  declare_model(
    N = N,
    U = rnorm(N),
    control_count = rbinom(N, size = 3, prob = 0.5),
    Y_star = rbinom(N, size = 1, prob = 0.3),
    W = case_when(Y_star == 0 ~ 0L,
                  Y_star == 1 ~ rbinom(N, size = 1, prob = proportion_hiding)),
    potential_outcomes(Y_list ~ Y_star * Z + control_count)
  ) +
  declare_inquiry(prevalence_rate = mean(Y_star)) +
  declare_assignment(Z = complete_ra(N)) + 
  declare_measurement(Y_list = reveal_outcomes(Y_list ~ Z),
                      Y_direct = Y_star - W) +
  declare_estimator(Y_list ~ Z, inquiry = "prevalence_rate", label = "list") + 
  declare_estimator(Y_direct ~ 1, inquiry = "prevalence_rate", label = "direct")

```

### Diagnosis

```{r, eval = FALSE}
declaration_17.4 |> 
  redesign(proportion_hiding = seq(from = 0, to = 0.3, by = 0.1), 
           N = seq(from = 500, to = 2500, by = 500)) |> 
  diagnose_design()
```

```{r, echo = FALSE, fig.height = 2.5, fig.width = 10}
if(run)
declaration_17.4 |> 
  redesign(proportion_hiding = seq(from = 0, to = 0.3, by = 0.1), 
           N = seq(from = 500, to = 2500, by = 500)) |> 
  diagnose_design() |>
  write_rds("saved/list2.rds")

read_rds("saved/list2.rds") |>
  tidy() |>  filter(diagnosand == "rmse") |>
  ggplot(aes(N, estimate, color = estimator)) + geom_line() + facet_grid(~proportion_hiding, labeller = label_both) + theme_bw()

```



### Negatively correlated items

* How would estimates be affected if the items selected for the list were *negatively* correlated?
* How would subject protection be affected?


### Negatively correlated items

```{r}
rho <- -.8 

correlated_lists <- 
  declare_model(
    N = 500,
    U = rnorm(N),
    control_1 = rbinom(N, size = 1, prob = 0.5),
    control_2 = correlate(given = control_1, rho = rho, draw_binary, prob = 0.5),
    control_count = control_1 + control_2,
    Y_star = rbinom(N, size = 1, prob = 0.3),
    potential_outcomes(Y_list ~ Y_star * Z + control_count)
  ) +
  declare_inquiry(prevalence_rate = mean(Y_star)) +
  declare_assignment(Z = complete_ra(N)) + 
  declare_measurement(Y_list = reveal_outcomes(Y_list ~ Z)) +
  declare_estimator(Y_list ~ Z) 
```

### Negatively correlated items

```{r}
draw_data(correlated_lists) |> ggplot(aes(control_count)) + 
  geom_histogram() + theme_bw()
```

### Negatively correlated items

```{r, eval = FALSE}
correlated_lists |> redesign(rho = c(-.8, 0, .8)) |> diagnose_design()
```

```{r, echo = FALSE, height = 2, width = 8}
if(run)
correlated_lists |> redesign(rho = c(-.8, 0, .8)) |> diagnose_design() |>
  write_rds("saved/correlated_list.rds")

read_rds("saved/correlated_list.rds") |> tidy() |> filter(diagnosand == "rmse") |> ggplot(aes(rho, estimate)) + geom_line() + ylim(0, NA) + theme_bw() + ylab("rmse")

```

* These trade-off against each other: the more accuracy you have the less protection you have

### Individual or group effects? {.smaller}

* This is typically used to estimate average *levels*

* However you *can* use it in the obvious way to get average levels for groups: this is equivalent to calculating group level heterogeneous effects

* Extending the idea you can even get *individual level* estimates: for instance you might use causal forests

* You can also use this to estimate the *effect* of an experimental treatment on an item that's measured using a list, without requiring individual level estimates:

$$Y_i = \beta_0 + \beta_1Z_i  + \beta_2Long_i + \beta_3Z_iLong_i$$



### Hiders and liars

* Note that here we looked at "hiders" -- people not answering the direct question truthfully

* See @li2019relaxing on bounds when the "no liars" assumption is threatened --- this is about whether people respond truthfully to the list experimental question


  

