---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
title: "Topics"
subtitle: "survey experiments"
author: "Macartan Humphreys"
bibliography: ../bib.bib
---

  
```{r, include = FALSE}

source("setup.R")

```


## List experiments {#survey-experiments}


* manipulations are used to *measure* things not to change things
* changing the wording changes the answer but this helps you learn about a stable quantity


### The list experiment: Motivation

* Multiple survey experimental designs have been generated to make it easier for subjects to answer **sensitive questions**

* The key idea is to use inference rather than measurement.

* Subjects are placed in different conditions and the conditions *affect* the **answers** that are given in such a way that you can infer some underlying quantity of interest

### The list experiment: Motivation

```{r, echo = FALSE, caption = "Questions and values collide in an answer"}
make_model("Z -> Question -> Answer <- Value") |> plot(textcol = "black", nodecol = "lightgrey")

```

This is an obvious DAG but the main point is to be clear that the *Value* is the quantity of interest and the value is not affected by the treatment, *Z*.

### The list experiment: Motivation

The list experiment supposes that:

1. Subjects do not want to give a direct answer to a question
2. They nevertheless are willing to truthfully answer an indirect question

**In other words**: sensitivities notwithstanding, *they are happy for the researcher to make correct inferences about them or their group*


### The list experiment: Strategy {.smaller}

* Respondents are given a short list and a long list.

* The long list differs from the short list in having *one extra item*---the sensitive item

* We ask how many items in each list does a respondent agree with: 

  * $Y_i(0)$ is the number of elements on a short list that a respondent agrees with
  * $Y_i(1)$ is the number of elements on a long list that a respondent agrees with
  * $Y_i(1) - Y_i(0)$ is an indicator for whether an individual agrees with the sensitive item
  * $\mathbb{E}[Y_i(1) - Y_i(0)]$ is the share of people agreeing with sensitive item



### The list experiment: Simplified example {.smaller}

How many of these do you agree with:

|        | Short list       | Long list          | "Effect"  |
|------|------------------|----------------|--------|
|        | "2 + 2 = 4"    | "2 + 2 = 4"                |   |
|        | "2 * 3 = 6"    | "2 * 3 = 6"  |   |
|        | "3 + 6 = 8"    | "Climate change is real" |   |
|        |            | "3 + 6 = 8"                |   |
| Answer | Y(0) = 2   | Y(1) = 4    |  Y(1) - Y(0) = 2 |


[Note: this is obviously not a good list. Why not?]

### The list experiment: Design

```{r}

declaration_17.3 <-
  declare_model(
    N = 500,
    control_count = rbinom(N, size = 3, prob = 0.5),
    Y_star = rbinom(N, size = 1, prob = 0.3),
    potential_outcomes(Y_list ~ Y_star * Z + control_count) 
  ) +
  declare_inquiry(prevalence_rate = mean(Y_star)) +
  declare_assignment(Z = complete_ra(N)) + 
  declare_measurement(Y_list = reveal_outcomes(Y_list ~ Z)) +
  declare_estimator(Y_list ~ Z, .method = difference_in_means, 
                    inquiry = "prevalence_rate")

diagnosands <- declare_diagnosands(
  bias = mean(estimate - estimand),
  mean_CI_width = mean(conf.high - conf.low)
)
```


### Diagnosis

```{r, eval = FALSE}
diagnose_design(declaration_17.3, diagnosands = diagnosands)
```

```{r, echo = FALSE}
if(run)
  diagnose_design(declaration_17.3, diagnosands = diagnosands) |>
  write_rds("saved/list1.rds")

read_rds("saved/list1.rds") |>
  reshape_diagnosis() |> select("Design", "Inquiry", "Bias", "Mean CI Width") |> kable()
  

```

### Tradeoffs: is the question really sensitive?


```{r}
declaration_17.4 <- 
  declare_model(
    N = N,
    U = rnorm(N),
    control_count = rbinom(N, size = 3, prob = 0.5),
    Y_star = rbinom(N, size = 1, prob = 0.3),
    W = case_when(Y_star == 0 ~ 0L,
                  Y_star == 1 ~ rbinom(N, size = 1, prob = proportion_hiding)),
    potential_outcomes(Y_list ~ Y_star * Z + control_count)
  ) +
  declare_inquiry(prevalence_rate = mean(Y_star)) +
  declare_assignment(Z = complete_ra(N)) + 
  declare_measurement(Y_list = reveal_outcomes(Y_list ~ Z),
                      Y_direct = Y_star - W) +
  declare_estimator(Y_list ~ Z, inquiry = "prevalence_rate", label = "list") + 
  declare_estimator(Y_direct ~ 1, inquiry = "prevalence_rate", label = "direct")

```

### Diagnosis

```{r, eval = FALSE}
declaration_17.4 |> 
  redesign(proportion_hiding = seq(from = 0, to = 0.3, by = 0.1), 
           N = seq(from = 500, to = 2500, by = 500)) |> 
  diagnose_design()
```

```{r, echo = FALSE, fig.height = 2.5, fig.width = 10}
if(run)
declaration_17.4 |> 
  redesign(proportion_hiding = seq(from = 0, to = 0.3, by = 0.1), 
           N = seq(from = 500, to = 2500, by = 500)) |> 
  diagnose_design() |>
  write_rds("saved/list2.rds")

read_rds("saved/list2.rds") |>
  tidy() |>  filter(diagnosand == "rmse") |>
  ggplot(aes(N, estimate, color = estimator)) + geom_line() + facet_grid(~proportion_hiding, labeller = label_both) + theme_bw()

```



### Negatively correlated items

* How would estimates be affected if the items selected for the list were *negatively* correlated?
* How would subject protection be affected?


### Negatively correlated items

```{r}
rho <- -.8 

correlated_lists <- 
  declare_model(
    N = 500,
    U = rnorm(N),
    control_1 = rbinom(N, size = 1, prob = 0.5),
    control_2 = correlate(given = control_1, rho = rho, draw_binary, prob = 0.5),
    control_count = control_1 + control_2,
    Y_star = rbinom(N, size = 1, prob = 0.3),
    potential_outcomes(Y_list ~ Y_star * Z + control_count)
  ) +
  declare_inquiry(prevalence_rate = mean(Y_star)) +
  declare_assignment(Z = complete_ra(N)) + 
  declare_measurement(Y_list = reveal_outcomes(Y_list ~ Z)) +
  declare_estimator(Y_list ~ Z) 
```

### Negatively correlated items

```{r}
draw_data(correlated_lists) |> ggplot(aes(control_count)) + 
  geom_histogram() + theme_bw()
```

### Negatively correlated items

```{r, eval = FALSE}
correlated_lists |> redesign(rho = c(-.8, 0, .8)) |> diagnose_design()
```

```{r, echo = FALSE, height = 2, width = 8}
if(run)
correlated_lists |> redesign(rho = c(-.8, 0, .8)) |> diagnose_design() |>
  write_rds("saved/correlated_list.rds")

read_rds("saved/correlated_list.rds") |> tidy() |> filter(diagnosand == "rmse") |> ggplot(aes(rho, estimate)) + geom_line() + ylim(0, NA) + theme_bw() + ylab("rmse")

```

* These trade-off against each other: the more accuracy you have the less protection you have

### Individual or group effects? {.smaller}

* This is typically used to estimate average *levels*

* However you *can* use it in the obvious way to get average levels for groups: this is equivalent to calculating group level heterogeneous effects

* Extending the idea you can even get *individual level* estimates: for instance you might use causal forests

* You can also use this to estimate the *effect* of an experimental treatment on an item that's measured using a list, without requiring individual level estimates:

$$Y_i = \beta_0 + \beta_1Z_i  + \beta_2Long_i + \beta_3Z_iLong_i$$



### Hiders and liars

* Note that here we looked at "hiders" -- people not answering the direct question truthfully

* See @li2019relaxing on bounds when the "no liars" assumption is threatened --- this is about whether people respond truthfully to the list experimental question


  

## Conjoint

### Conjoints

* Conjoints are factorial experiments in which you vary multiple features of a description at once
* You can then define the effect of a given feature on responses conditional on the distribution of other features
* The *average* effect of a feature is taken both with respect to subjects and with respect to background profiles

* In this literature the estimand is often called the "Average Marginal Component Effect" but it generally corresponds to the ATE from a factorial as described by Fisher
* For simple random assignments you can use differences in means; for blocked assignments you can take account of blocks


### The strengths of a conjoint  {.smaller}

* You can vary many features simultaneously
* You may get more natural descriptions of more fully fleshed out options
* You may avoid desirability biases

See @hainmueller2014causal

### Pitfalls  {.smaller}

* The use of causal inference techniques can mislead into thinking that the effects of attributes themselves are being studied; need to remain conscious that these are focused on seeing how changing questions change answers --- it is the effect of *descriptions* on *responses* not the effect of *features* on *actions* --- since ultimately you are asking for hypothetical evaluations these might be best thought of as a nice tool to measure (stated) preferences
* You generally end up *controlling* for various features,  rather then having them vary naturally; so, for instance, if you have both origin and education you are not getting effects of origin that work via education (or beliefs about education that are due to beliefs about origin)


### Conjoint forced choice  {.smaller}

Often the focus is on the effect of described features on the indicated preference for one candidate over another.

Imagine this setup:

Two candidates. 

Two attributes:

- Corrupt (C) vs Non-corrupt (N)  
- Left-handed (L) vs Right-handed (R)

Decision rule:

1. Voters always choose a **non-corrupt candidate over a corrupt one**.
2. If both are corrupt or both are non-corrupt, they prefer **right-handed**.
3. Otherwise, they flip a coin.

What is the average effect of corruption on vote choice? What is the average effect of right handedness?

### Effects across all profiles  {.smaller}

Define $Y(C_A, R_A, C_B, R_B)$ as the decision to vote for candidate A given the different values of corrutopn and handedness for A and B.

Then the average effect of corruption on the vote for 1 for a given person, across possible contests, is:

$$\frac{1}8 \sum_{r_a=0}^1 \sum_{c_b=0}^1 \sum_{r_b=0}^1 \left(Y(1, r_a, c_b, r_b) - Y(0, r_a, c_b, r_b)\right)$$

This estimand is defined weighting all profiles equally. But it is up to you how you want to weight.

### Effect of corruption (C vs N) {.smaller}

We average over all distinct contexts:

| A (handedness) | B  | Effect of C on selecting A |
|----------------|----|----------------------------|
| L | CL | -0.5 |
| R | CL | 0 |
| L | CR | -1 |
| R | CR | -0.5 |
| L | NL | -0.5 |
| R | NL | -1 |
| L | NR | 0 |
| R | NR | -0.5 |
|   |    | **Average = -0.5** |


### Effect of right-handedness (R vs L) {.smaller}

We average over all distinct contexts:

| A (corruption) | B  | Effect of R on selecting A |
|----------------|----|----------------------------|
| C | CL | 0.5 |
| N | CL | 0 |
| C | CR | 0.5 |
| N | CR | 0 |
| C | NL | 0 |
| N | NL | 0.5 |
| C | NR | 0 |
| N | NR | 0.5 |
|   |    | **Average = 0.25** |

### Effects across all profiles {.smaller}

This average effect is the ACME.

Intuitively:

* Non corruption: 1/8 of the time A will win regardless of corruption. 1/8 A will lose regardless of corruption; 2/8 of the time corruption takes A from a sure win to a sure loss; 4/8 of the time corruption gets you to or from a tie.

* Righthandedness: 1/2 of the time handedness is irrelevant. 1/2 the time there are ties and handedness gets you to or from a tie.

* Example really highlights how the averaging is over conditions (over races). If you are interested in the role of corruption in a particular type of race (or a target race) this is not the quantity you want

### Do it with Lin {.smaller}

```{r}

Y <- function(cA, rA, cB, rB) 
  case_when((cA < cB) | ((cA == cB) & (rA > rB))  ~ 1,
            (cA > cB) | ((cA == cB) & (rA < rB))  ~ 0,
            .default = .5)

design <-
  
  declare_model(
    data = expand_grid(cA = 0:1, rA = 0:1, cB = 0:1, rB = 0:1),
    Y = Y(cA, rA, cB, rB)) +
  
  declare_inquiry(
    tau_C = mean(Y(1, rA, cB, rB) - Y(0, rA, cB, rB)),
    tau_R = mean(Y(cA, 1, cB, rB) - Y(cA, 0, cB, rB))) +
  
  declare_estimator(Y ~ cA, covariates = ~rA + cB + rB,  
                    .method = lm_lin, inquiry = "tau_C", label = "C") +
  
  declare_estimator(Y ~ rA, covariates = ~cA + cB + rB,  
                    .method = lm_lin, inquiry = "tau_R", label = "R")
```

### Do it with Lin {.smaller}

```{r}
run_design(design) |> select(inquiry, estimand, estimate) |>  kable()
```
