---
title: "Topics"
subtitle: "Synthetic controls"
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
    link-color: orange
author: "Macartan Humphreys"
bibliography: ../bib.bib
---

```{r synthset, include = FALSE}
if(exists("child")) source(here::here("slides", "setup.R"))
if(!exists("child")) source("../setup.R")

library(SCtools)
library(Synth)
library(tidysynth)
# devtools::install_github("edunford/tidysynth")

require(tidysynth)

```

## Synthetic controls

Interested in the effect of an intervention on a single unit.

Available data: time series for unit of interest and possible comparison groups 


### Core idea {.smaller}

* Generate a Frankenstein control from weighting a set of control units so that they collectively track the unit of interest and can  plausibly serve as a counterfactual.

* Identify weights $w$ that minimize $\left(Y_i^s(0) - \sum_{j\neq i} w_j Y_j^s(0)\right)^2$ in time period(s) $s$, prior to treatment; use these to predict counterfactual outcomes $Y^t_i(0)$ in later periods, $t$


* Note, in @abadie2003economic weights are chosen to minimize $(X_1 - X_0W)'V(X_1 - X_0W)$ where $X_1$, $X_0$ are covariates in the unit of interest and the "donor" units respectively. $V$ is chosen so that the pre treatment *outcomes* are as close as possible. 


### Optimization problem

In practice weights are found using a two stage minimization procedure:

1. Given matrix $V$ choose $W(V)$ to minimize $(X_1-X_0W)'V(X1-X_0W)$
2. Choose $V$ to minimize  $(Z_1-Z_0W(V))'(X1-X_0W(V))$

So in effect you end up with weights on cases and also weights on variables

### Basque example

* The seminal study is: @abadie2003economic which examines the impact of conflict on the Basque area

* A vignette showing how to use the [`SCtools`](https://cran.r-project.org/web/packages/SCtools/index.html)   package to implement synthetic controls for this example is (here)[https://cran.r-project.org/web/packages/SCtools/vignettes/replicating-basque.html]


*  We will walk through this and then interrogate this with a common package and a tidy version

### Basque example {.smaller}

Packages and data:

```{r, message = FALSE}
library(SCtools)
library(Synth)
data("basque")
```

### Basque example {.smaller}


Some fairly nasty data manipulation:

```{r}
f_dataprep  <- function(data, time_plot = min(data$year):max(data$year))
  dataprep(
  foo = data,
  predictors = c("school.illit", "school.prim", "school.med",
     "school.high", "school.post.high", 
    "invest"),
  predictors.op = "mean",
  time.predictors.prior = 1964:1969,
  special.predictors = list(
    list("gdpcap", 1960:1969 ,"mean"),
    list("sec.agriculture", seq(1961, 1969, 2), "mean"),
    list("sec.energy", seq(1961, 1969, 2), "mean"),
    list("sec.industry", seq(1961, 1969, 2), "mean"),
    list("sec.construction", seq(1961, 1969, 2), "mean"),
    list("sec.services.venta", seq(1961, 1969, 2), "mean"),
    list("sec.services.nonventa", seq(1961, 1969, 2), "mean"),
    list("popdens",               1969,               "mean")),
  dependent = "gdpcap",
  unit.variable = "regionno",
  unit.names.variable = "regionname",
  time.variable = "year",
  treatment.identifier = 17,
  controls.identifier = c(2:16, 18),
  time.optimize.ssr = 1960:1969,
  time.plot = time_plot)

dataprep.out  <- f_dataprep(basque)
```



### Implementation (with messages) {.smaller}

```{r}
synth.out <- synth(data.prep.obj = dataprep.out, method = "BFGS")
```

### Results {.smaller}

This produces the following weights, selected to render the pre-treatment trends of the synthetic and actual controls as similar as possible. 


```{r, echo = FALSE}

basque |> group_by(regionno) |> slice(1) |>  select(regionno, regionname) |>
  left_join(data.frame(
    regionno = rownames(synth.out$solution.w ) |> as.numeric(),
    weight =  synth.out$solution.w)) |>
  kable(digits = 2)


```

### Plotting the counterfactual

Lets plot by hand and compare with package

```{r}

basque_df <- basque |> filter(regionno == 17) |> 
  mutate(counterfactual = .15*basque$gdpcap[basque$regionno == 14] + .85*basque$gdpcap[basque$regionno == 10] )

basque_df |> ggplot(aes(year, gdpcap)) + geom_line(color = "blue") + 
  geom_line(aes(year, counterfactual), color = "red") + theme_bw() +
  ylab("real per-capita GDP (1986 USD, thousand)") + xlab("year")


```


### Plotting 2

```{r}
path.plot(synth.res = synth.out, dataprep.res = dataprep.out,
          Ylab = "real per-capita GDP (1986 USD, thousand)", Xlab = "year",
          Ylim = c(0, 12), Legend = c("Basque country",
                                      "synthetic Basque country"), 
          Legend.position = "bottomright")
```


### Then what?

What is your estimate?

How do you do inference?


* Estimate could be at a particular point in time or averaging over a range. Inference often based on placebos: assess  whether estimated effects are large relative to what you would get if you looked for effects in random places. 

*  This has an RI flavor to it, but we are not exploiting any actual randomization.

* Alternatives are to do sensitivity analyses, bootstrap (though this is really capturing sampling variability), and perhaps using estimates of the uncertainty from errors in pre treatment predictions.

### `tidysynth` implementation {.smaller}

`tidysynth` a wrapper for `synth` and tidier for sure

Vignette on: Impact of Proposition 99 on cigarette consumption in California. https://github.com/edunford/tidysynth


```{r}
data("smoking")
smoking |> head() |> kable()
```

### `tidysynth` implementation

In a pipe:

```{r}
smoking_out <-
  
  smoking %>%
  
  # initial the synthetic control object
  synthetic_control(
    outcome = cigsale, # outcome
    unit = state, # unit index in the panel data
    time = year, # time index in the panel data
    i_unit = "California", # unit where the intervention occurred
    i_time = 1988, # time period when the intervention occurred
    generate_placebos=T # generate placebo synthetic controls (for inference)
    )

```

### add predictor information {.smaller}

```{r}

smoking_out <- smoking_out %>%
  
  # Generate the aggregate predictors used to fit the weights
  
  # average log income, retail price of cigarettes, and proportion of the
  # population between 15 and 24 years of age from 1980 - 1988
  generate_predictor(time_window = 1980:1988,
                     ln_income = mean(lnincome, na.rm = T),
                     ret_price = mean(retprice, na.rm = T),
                     youth = mean(age15to24, na.rm = T)) %>%
  
  # average beer consumption in the donor pool from 1984 - 1988
  generate_predictor(time_window = 1984:1988, 
                     beer_sales = mean(beer, na.rm = T)) %>%
  
  # Lagged cigarette sales 
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) %>%
  generate_predictor(time_window = 1980, cigsale_1980 = cigsale) %>%
  generate_predictor(time_window = 1988, cigsale_1988 = cigsale) 

```

### implement  {.smaller}

```{r}

smoking_out <- smoking_out %>%
  
  
  # Generate the fitted weights for the synthetic control
  generate_weights(optimization_window = 1970:1988, # time for optimization
                   margin_ipop = .02,sigf_ipop = 7,bound_ipop = 6 # options
  ) %>%
  
  # Generate the synthetic control
  generate_control()


```

### plot trends

```{r}

smoking_out %>% plot_trends()


```

### plot weights

```{r}
smoking_out %>% plot_weights()


```

### plot placebos

```{r}
smoking_out %>% plot_placebos()
```


### Design declaration {.smaller}

```{r}

sc_helper <- function(data, treatment_time = 2010)

  data |>
  
  # initial the synthetic control object
  synthetic_control(
    outcome = Y, # outcome
    unit = state, # unit index in the panel data
    time = year, # time index in the panel data
    i_unit = "01", # unit where the intervention occurred
    i_time = treatment_time, # time period when the intervention occurred
    generate_placebos = F  # generate placebo synthetic controls
    )  |>
  
  generate_predictor(time_window = 2000:2020,
                     lnincome = mean(ln_income, na.rm = T),
                     youth = mean(age15to24, na.rm = T)) %>%
  generate_weights(optimization_window = 2000:2010, # time for optimization
                   margin_ipop = .02,sigf_ipop = 7,bound_ipop = 6 # options
  ) |>
  generate_control()
```


### Declare an estimator

```{r}

sc_estimator <- function(data, treatment_time = 2010) 
  
  data.frame(estimate = 
               sc_helper(data, treatment_time = 2010) |>
               grab_synthetic_control() |> 
               mutate(difference = real_y - synth_y) |>
               filter(time_unit >= treatment_time) |> 
               pull(difference) |> 
               mean()
)

```

### Design

```{r}

design <-
  declare_model(
    state = add_level(50, 
                      youth_mean = runif(N, .2, .7), 
                      income_mean = runif(N, 1, 5), 
                      b = runif(N, 0, 10)),
    time = add_level(21, year = 2000:2020, nest = FALSE),
    unit = cross_levels(join_using(state, time))) +
  
  declare_model(
    treatment = 1*(state == "01" & (year >=2010)),
    age15to24 = youth_mean + .5*runif(N, -.1, .1) - .2*(year- 2000),
    ln_income = income_mean + .5*rlnorm(N),
    Y = 5*age15to24 + .1*(year- 2000)^2 +  1*ln_income + b* treatment) +
  
  declare_inquiry(b[1]) +
  
  declare_estimator(handler = label_estimator(sc_estimator))

```

### Inspection

```{r}

draw_data(design) |> sc_helper()  |> plot_trends()

```

### Simulation

```{r}
simulations <- simulate_design(design, sims = 10)

simulations |> ggplot(aes(estimand, estimate)) + geom_point()

```
