---
title: "Topics"
subtitle: "Synthetic controls"
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
    link-color: orange
author: "Macartan Humphreys"
bibliography: ../bib.bib
---

```{r synthset, include = FALSE}
if(exists("child")) source("setup.R")
if(!exists("child")) source("../setup.R")

library(SCtools)
library(Synth)
```

## Synthetic controls

Interested in the effect of an intervention on a single unit.

Available data: time series for unit of interest; 


### Core idea

Generate a frankenstein control from weighting a set of control outcomes so that they collectively track the unit of interest and can  plausibly serve as a counterfactual.

```{r, eval = FALSE, include = FALSE}
library(nloptr)


find_weights <- function(data, cutoff) {
  Y <- data$Y[data$T < cutoff]
  Ys <- as.matrix(data[data$T < cutoff, c("YA", "YB", "YC")])
  
  obj_fun <- function(w)  sum((Y - Ys %*% w)^2)
  
  w0 <- rep(1 / ncol(Ys), ncol(Ys))
  
  w <- 
    nloptr(
    x0 = w0, 
    eval_f = obj_fun,
    eval_g_eq = function(w) sum(w) - 1, 
    lb = rep(0, ncol(Ys)),
    opts = list(algorithm = "NLOPT_LN_COBYLA", xtol_rel = 1e-6)
  )$solution
  
  print(w)
  
  data |>
    mutate(Y0 = w[1]*data$YA + w[2]*data$YB + w[3]*data$YC)
}



design <-
  declare_model(N  = 10,
                T = 1:N,
                X = runif(N),
                YA = 1 + .25*T + .3*X,
                YB = 0 + .5*T + .2*X,
                YC = 1 + .3*Y + .4*X,
                Y = 2 + .4*X + .3*X*(T>5)

                )

data <- design()

find_weights(data, 5) |>
  gather(var, val, -ID, -T, -X) |>
  ggplot(aes(T, val, color = var)) + geom_line()

```



### Basque example

```{r, echo = FALSE, eval = FALSE}
data("basque")

dataprep.out  <- 
  dataprep.out <- dataprep(
  foo = basque,
  predictors = c("school.illit", "school.prim", "school.med",
    "school.high", "school.post.high", "invest"),
  predictors.op = "mean",
  time.predictors.prior = 1964:1969,
  special.predictors = list(
    list("gdpcap", 1960:1969 ,"mean"),
    list("sec.agriculture", seq(1961, 1969, 2), "mean"),
    list("sec.energy", seq(1961, 1969, 2), "mean"),
    list("sec.industry", seq(1961, 1969, 2), "mean"),
    list("sec.construction", seq(1961, 1969, 2), "mean"),
    list("sec.services.venta", seq(1961, 1969, 2), "mean"),
    list("sec.services.nonventa", seq(1961, 1969, 2), "mean"),
    list("popdens",               1969,               "mean")),
  dependent = "gdpcap",
  unit.variable = "regionno",
  unit.names.variable = "regionname",
  time.variable = "year",
  treatment.identifier = 17,
  controls.identifier = c(2:16, 18),
  time.optimize.ssr = 1960:1969,
  time.plot = 1955:1997)
```


## In practice

The [SCtools](https://cran.r-project.org/web/packages/SCtools/index.html) is useful package.

We will go through their [vignette on the Basque conflict](https://cran.r-project.org/web/packages/SCtools/vignettes/replicating-basque.html) 

