---
title: 'Design and analysis of experiments'
subtitle: HU | Spring 2026 (Winter term 25-26)
abstract: "The course addresses the design and analysis of experiments. It is a hands-on course in which theoretical results are introduced through lecture, demonstrated in practice, and worked on in groups. Topics include randomization schemes, analysis strategies, design evaluation, and workflows."
author: "Macartan Humphreys"
bibliography: ../slides/bib.bib
format: 
  pdf:
    number-sections: true
    # Adjust text size and margins
    documentclass: extarticle
    classoption: a4paper,10pt
    margin-left: 1in
    margin-right: 1in
    margin-top: 0.8in
    margin-bottom: 0.8in
urlcolor: orange
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Class resources

* All resources: [https://macartan.github.io/ci](https://macartan.github.io/ci)

* This syllabus: [https://macartan.github.io/ci/experiments_syllabus_2026.pdf](https://macartan.github.io/ci/experiments_syllabus_2026.pdf)

* Slides: [https://macartan.github.io/ci/experiments_2026.html](https://macartan.github.io/ci/experiments_2026.html)

* [Student survey](https://cloud.wzb.eu/apps/forms/s/8QmokT5GeQfkmkrBHDzrzN4j)

* Git repo [https://github.com/macartan/ci](https://github.com/macartan/ci): This repo is for both my causal inference class and this experiments class.

# Times and locations 

Meetings will be held at the WZB

| Date     | Time    | Location     | Topic                          |
|----------|---------|--------------|--------------------------------|
| 9 Jan 2026    | 9 - 13 | WZB A 305  | Getting started |
| 16 Jan  2026  | 9 - 13   | WZB A 305  | Doing experiments |
| 30 Jan 2026    | 9 - 13   | WZB A 305  | Causality and causal inquiries |
| 6 Feb 2026    | 9 - 13   | WZB A 305  | Answer strategies |
| 20 Feb 2026   | 9 - 13   | WZB A 305  | Data strategies and design evaluation |
| 27 Feb 2026   | 9 - 13   | WZB B 001  | Topics: downstream experimentation, spillover experiments, patient preference trials, survey experiments |


The sessions will be structured roughly as follows.

* 9:00 - 11:00: lecture or discussions of readings
* 11:00 - 11:15: break
* 11:15 - 13:00: discussions, team presentations and in-class exercises





# Expectations

## Required

* Do the readings and take part in discussion  (25%)
* Take part in field experiment I  (25%)
* Design an experiment as a candidate for survey experiment II  (25%)
* Short in class exam   (25%)


## Optional
* Prepare a research design or short paper, perhaps building on existing work. Typically this contains: 
  - a problem statement
  - a description of a method to address the problem
  - analytic or simulation based results describing properties of the solution
  - a discussion of implications for practice. 
  - A passing paper will illustrate subtle features of a method; a good paper will identify unknown properties of a method; an excellent paper will develop a new method. 






# Sessions and readings

The course combines lectures, discussion, and activities. 

All readings are linked from this document. 

I will draw material especially from @blair2023research: [Research Design in the Social Sciences](https://book.declaredesign.org/)

Recommended but not open access: [@gerber2012field](https://wwnorton.com/books/9780393979954) with supplementary material available here: [https://isps.yale.edu/FEDAI](https://isps.yale.edu/FEDAI)

Readings for each session are given below.

## Intro

Aims: 

* Introduce the class
* Practice running experiments by hand
* Introduce and discuss field [experiment I](in_class_experiment.html)
* To prepare for next session: proposed adaptation of field experiment 1

**Course outline, tools** 



## Experimentation

Aims:

* Review lifecycle of an experiment (discussion)
* Discuss ethical challenges to experimentation
* Develop group field experiment 1

Required reading:

* [The Hijab penalty](https://onlinelibrary.wiley.com/doi/full/10.1111/ajps.12627) and the  [Appendix](https://onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1111%2Fajps.12627&file=ajps12627-sup-0001-Appendix.pdf)
* [Johnson: Protecting the Community: Lessons from the Montana Flyer Project](https://www.cambridge.org/core/journals/ps-political-science-and-politics/article/protecting-the-community-lessons-from-the-montana-flyer-project/2595ED0DB8189897DF6BDB121BBB83CA)
* [DD, Ch 2](https://book.declaredesign.org/introduction/what-is-a-research-design.html) introduces ideas at a high level
* [Design lifecycle](https://book.declaredesign.org/lifecycle) @blair2023research p 319 - 358
* [@humphreys2015reflections](https://macartan.github.io/assets/pdf/papers/2015_reflections_ethics.pdf) on ethics in experiments

Optional reading:

* [@alvarez2018research](https://www.researchgate.net/publication/324240650_Research_Replication_Practical_Considerations/link/5ca3688b458515f7851d7b2e/download?)
* [@humphreys2013fishing](https://macartan.github.io/assets/pdf/papers/2013_fishing.pdf) on fishing and registration


## Causality

Aims: 

* Understand the potential outcomes framework
* Understand identification challenges and the role of randomization
* Understand varieties of estimands

Required:

* [Holland](https://www.jstor.org/stable/2289064) is a beautiful classic reading on the fundamental problem of causal inference
* [DD, Ch 7](https://book.declaredesign.org/declaration-diagnosis-redesign/defining-inquiry.html)

Optional:

* [II Ch 2](https://integrated-inferences.github.io/book/02-causal-models.html), goes over both potential outcomes and DAGs
* [@keele2015statistics](https://hrr.w.uib.no/files/2019/01/Keele_2015_causal.pdf) gives a good high level overview of many key ideas in this course
* [@imai2008misunderstandings](https://gking.harvard.edu/files/matchse.pdf) 
* [@hernanwhatif](https://miguelhernan.org/whatifbook)  (Section 3.1, 7.2, 8.4, 10.1)
* [@hernanwhatif](https://miguelhernan.org/whatifbook) Ch 6
* [II Ch 4](https://macartan.github.io/integrated_inferences/HJC4.html)
* @lundberg2021your

## Answer strategies

Aims:

* Understand alternative strategies to estimating treatment effects
* Understand design based strategies for estimating standard errors and $p$-values
* Understand confidence intervals
* Understand the role of covariates in experimental design

Required:

* [@freedman2008regression](https://www.stat.berkeley.edu/~census/neyregr.pdf) helps make connections between design based inference and regression 
* [@lin2012agnostic](https://arxiv.org/pdf/1208.2301.pdf) relieves some worries you might have after reading @freedman2008regression

Optional:

* [@bind2020possible](https://scholar.harvard.edu/files/marie-abele/files/pnas.pdf)  on RI
* [@chang2024exact](https://onlinelibrary.wiley.com/doi/full/10.3982/ECTA20289) follows up on @lin2012agnostic.


## Experimental Design and Evaluation

Aims:

* Understand alternative assignment strategies
* Understand statistical power
* Understand how to assess alternative diagnosands

Required:

* [DD, Ch 7](https://book.declaredesign.org/declaration-diagnosis-redesign/diagnosing-designs.html)
* [DD, Ch 8](https://book.declaredesign.org/declaration-diagnosis-redesign/crafting-data-strategy.html) on data strategies
* [DD, Ch 9](https://book.declaredesign.org/declaration-diagnosis-redesign/choosing-answer-strategy.html) on answer strategies
* [DD, Ch 13](https://book.declaredesign.org/declaration-diagnosis-redesign/declaration-in-code.html) gives more practical guidance on getting started

Optional:

## Topics

Aims:

* Topics TBD

Required reading (TBD):

### downstream experimentation

### spillover experiments

* [@aronow2017estimating](https://arxiv.org/abs/1305.6156) on spillovers

### patient preference trials

### survey experiments

### mediation experiments

* [@imai2010general](https://imai.fas.harvard.edu/research/files/BaronKenny.pdf) on mediation


\newpage


# Prerequisites and tools

You should already have background in statistics up to the point of feeling comfortable with regression. 

In addition you should know some `R`. Really, the more you can invest on getting on top of `R` before the class the better.  


Before the first class please make sure your `R` is up-to-date and that you are working in `R studio`. Then make sure you have the following packages installed.

```{r, eval = FALSE}
pacman::p_load(
  rstan, 
  dagitty, 
  DeclareDesign, 
  CausalQueries,
  tidyverse,
  knitr
  )
```

Resources:

* Resources for learning `R`: http://www.r-bloggers.com/how-to-learn-r-2/
* Hadley Wickham's [R for Data Science](https://r4ds.hadley.nz/)
* [Grant McDermott on data science](https://grantmcdermott.com/ds4e)


## Writing with  `qmd`

Please plan to do drafting in `Quarto`. This is a simple markup language that lets you integrate writing and coding. This document is written in `Quarto` and the slides will be also. 

The key thing is that you can insert code chunks like this.

```{r}
# Define a random number
x <- rnorm(1)
```


Code like this is run as the document compiles, and results can be accessed as needed, like this: we just sampled the random number $x =$ `r x`.

* See [@wickham2023r](https://r4ds.hadley.nz/) on [Workflow](https://r4ds.hadley.nz/workflow-scripts) and [Quarto](https://r4ds.hadley.nz/quarto)


I recommend using `Rstudio` as an editor. More information here: 
https://quarto.org/docs/get-started/hello/rstudio.html



## AI

AI can be a great aid to help understand this material and also to develop code. However it should only be used as an aid: it makes many many mistakes; you remain responsible for your code and more importantly for your comprehension.


# References


```{r, eval = FALSE}
Survey Experiments
Mullinix, K. J., Leeper, T. J., Druckman, J. N., & Freese, J. (2015). "The Generalizability of Survey Experiments." Journal of Experimental Political Science, 2(2), 109-138.
Link: https://doi.org/10.1017/XPS.2015.19 (Open access via Cambridge Core)
Why: Discusses external validity and generalizability of survey experiments
Krupnikov, Y., & Levine, A. S. (2014). "Cross-Sample Comparisons and External Validity." Journal of Experimental Political Science, 1(1), 59-80.
Link: https://doi.org/10.1017/XPS.2014.6 (Open access via Cambridge Core)
Why: Addresses external validity across samples
Coppock, A., & McClellan, O. A. (2019). "Validating the Demographic, Political, Psychological, and Experimental Results Obtained from a New Source of Online Survey Respondents." Research & Politics, 6(2).
Link: https://doi.org/10.1177/2053168018822174 (Open access)
Why: Discusses online survey platforms and validity
Mutz, D. C. (2011). "Population-Based Survey Experiments." Princeton University Press (selected chapters often available open access)
Why: Overview of survey experiments in political science

Spillover Experiments
Bowers, J., Fredrickson, M. M., & Panagopoulos, C. (2013). "Reasoning about Interference Between Units: A General Framework." Political Analysis, 21(1), 97-124.
Link: https://www.jstor.org/stable/23359636 (JSTOR - most libraries)
Why: Framework for thinking about spillovers
Sinclair, B., McConnell, M., & Green, D. P. (2012). "Detecting Spillover Effects: Design and Analysis of Multilevel Experiments." American Journal of Political Science, 56(4), 1055-1069.
Link: https://www.jstor.org/stable/23359636 (JSTOR - most libraries)
Why: Methods for detecting spillovers in field experiments

Mediation Experiments
Imai, K., Keele, L., & Tingley, D. (2010). "A General Approach to Causal Mediation Analysis." Psychological Methods, 15(4), 309-334.
Link: https://doi.org/10.1037/a0020761 (Often available open access)
Why: General framework for mediation (already in your list)
Acharya, A., Blackwell, M., & Sen, M. (2016). "Explaining Causal Findings Without Bias: Detecting and Assessing Direct Effects." American Political Science Review, 110(3), 512-529.
Link: https://www.jstor.org/stable/24806015 (JSTOR - most libraries)
Why: Discusses direct effects and mediation in experiments

Downstream Experimentation
Gerber, A. S., & Green, D. P. (2017). "Field Experiments on Voter Mobilization: An Overview of a Burgeoning Literature." In Handbook of Experimental Economics and Political Science (often available through libraries)
Why: Overview of follow-up and downstream effects
Green, D. P., & Gerber, A. S. (2019). "Get Out the Vote: How to Increase Voter Turnout." Brookings Institution Press (supplementary materials often open access)
Why: Discusses long-term effects of voter mobilization experiments
Patient Preference Trials / Choice Experiments
Hainmueller, J., Hopkins, D. J., & Yamamoto, T. (2014). "Causal Inference in Conjoint Analysis: Understanding Multidimensional Choices via Stated Preference Experiments." Political Analysis, 22(1), 1-30.
Link: https://www.jstor.org/stable/24572619 (JSTOR - most libraries)
Why: Conjoint analysis for political preferences
Bansak, K., Hainmueller, J., Hopkins, D. J., & Yamamoto, T. (2021). "Beyond the Breaking Point? Survey Satisficing in Conjoint Experiments." Political Science Research and Methods, 9(1), 53-71.
Link: https://doi.org/10.1017/psrm.2019.28 (Open access via Cambridge Core)
Why: Methodological considerations in choice experiments

General Experimental Design Discussion Pieces
Gerber, A. S., & Green, D. P. (2012). "Field Experiments: Design, Analysis, and Interpretation." Norton (supplementary materials: https://isps.yale.edu/FEDAI - open access)
Why: Comprehensive overview with open access supplementary materials
Druckman, J. N., Green, D. P., Kuklinski, J. H., & Lupia, A. (2011). "Experiments: An Introduction to Core Concepts." In Cambridge Handbook of Experimental Political Science (many chapters available through libraries)
Why: Core concepts in experimental political science
Morton, R. B., & Williams, K. C. (2010). "Experimental Political Science and the Study of Causality: From Nature to the Lab." Cambridge University Press (often available through libraries)
Why: Philosophical and methodological foundations
Power Analysis & Design Evaluation
Cohen, J. (1988). "Statistical Power Analysis for the Behavioral Sciences." Routledge (classic reference, often available through libraries)
Why: Standard reference on power analysis
Lakens, D. (2022). "Sample Size Justification." Collabra: Psychology, 8(1).
Link: https://doi.org/10.1525/collabra.33267 (Open access)
Why: Modern perspective on sample size justification

Ethics & Field Experiments
Desposato, S. (2016). "Ethics and Experiments: Problems and Solutions for Social Scientists and Policy Professionals." Routledge (some chapters available through libraries)
Why: Ethics in field experiments
Michelitch, K. (2015). "Detecting Election Fraud from Irregularities in Vote-Share Distributions." Political Analysis, 23(1), 1-24.
Link: https://www.jstor.org/stable/24572619 (JSTOR - most libraries)
Why: Raises ethical and methodological questions

Randomization & Analysis
Gerber, A. S., & Green, D. P. (2000). "The Effects of Canvassing, Telephone Calls, and Direct Mail on Voter Turnout: A Field Experiment." American Political Science Review, 94(3), 653-663.
Link: https://www.jstor.org/stable/2585837 (JSTOR - most libraries)
Why: Classic field experiment, good for discussion

```  