---
title: 'Design and analysis of experiments'
subtitle: HU | Spring 2026 (Winter term 25-26)
abstract: "The course addresses the design and analysis of experiments. It is a hands-on course in which theoretical results are introduced through lecture, demonstrated in practice, and worked on in groups. Topics include randomization schemes, analysis strategies, design evaluation, and workflows."
author: "Macartan Humphreys"
bibliography: ../slides/bib.bib
format: 
  pdf:
    number-sections: true
    # Adjust text size and margins
    documentclass: extarticle
    classoption: a4paper,10pt
    margin-left: 1in
    margin-right: 1in
    margin-top: 0.8in
    margin-bottom: 0.8in
urlcolor: orange
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Class resources

-   All resources: <https://macartan.github.io/ci>

-   This syllabus: <https://macartan.github.io/ci/experiments_syllabus_2026.pdf>

-   Slides: <https://macartan.github.io/ci/experiments_2026.html>

-   [Student survey](https://cloud.wzb.eu/apps/forms/s/8QmokT5GeQfkmkrBHDzrzN4j)

-   Git repo <https://github.com/macartan/ci>: This repo is for both my causal inference class and this experiments class.

# Times and locations

Meetings will be held at the WZB

| Date | Time | Location | Topic |
|--------------|--------------|--------------|------------------------------|
| 9 Jan 2026 | 9 - 13 | WZB A 305 | Getting started |
| 16 Jan 2026 | 9 - 13 | WZB A 305 | Doing experiments |
| 30 Jan 2026 | 9 - 13 | WZB A 305 | Causality and causal inquiries |
| 6 Feb 2026 | 9 - 13 | WZB A 305 | Answer strategies |
| 20 Feb 2026 | 9 - 13 | WZB A 305 | Data strategies and design evaluation |
| 27 Feb 2026 | 9 - 13 | WZB B 001 | Topics: downstream experimentation, spillover experiments, patient preference trials, survey experiments |

The sessions will be structured roughly as follows.

-   9:00 - 11:00: lecture or discussions of readings
-   11:00 - 11:15: break
-   11:15 - 13:00: discussions, team presentations and in-class exercises

# Expectations

## Required

-   Do the readings and take part in discussion (30%)
-   Take part in field experiment I (30%)
-   Design an experiment as a candidate for survey experiment II (40%)

## Optional

-   Prepare a research design or short paper, perhaps building on existing work. Typically this contains:
    -   a problem statement
    -   a description of a method to address the problem
    -   analytic or simulation based results describing properties of the solution
    -   a discussion of implications for practice.
    -   A passing paper will illustrate subtle features of a method; a good paper will identify unknown properties of a method; an excellent paper will develop a new method.

# Sessions and readings

The course combines lectures, discussion, and activities.

All readings are linked from this document.

I will draw material especially from @blair2023research: [Research Design in the Social Sciences](https://book.declaredesign.org/)

Recommended but not open access: [\@gerber2012field](https://wwnorton.com/books/9780393979954) with supplementary material available here: <https://isps.yale.edu/FEDAI>

Readings for each session are given below.

## Intro

Aims:

-   Introduce the class
-   Practice running experiments by hand
-   Introduce and discuss field [experiment I](in_class_experiment.html)
-   To prepare for next session: proposed adaptation of field experiment 1

**Course outline, tools**

## Experimentation

Aims:

-   Review lifecycle of an experiment (discussion)
-   Discuss ethical challenges to experimentation
-   Discuss a set of strong experiments
-   Develop group field experiment 1 and decide on strategy

**To do before class**: 1. the readings; 2. develop specific proposals in small groups for field experiment 1 (4 slides), completing individually or collectively a copy of the [EGAP research design form](https://egap.github.io/learningdays-resources/Exercises/design-form.html)

Required reading:

on ethics:

-   [Johnson: Protecting the Community: Lessons from the Montana Flyer Project](https://www.cambridge.org/core/journals/ps-political-science-and-politics/article/protecting-the-community-lessons-from-the-montana-flyer-project/2595ED0DB8189897DF6BDB121BBB83CA)
-   [@humphreys2015reflections](https://macartan.github.io/assets/pdf/papers/2015_reflections_ethics.pdf) on ethics in experiments

some cases:

-   Example 1: @gerber2008social
-   Example 2: @lowe2021types
-   [The Hijab penalty](https://onlinelibrary.wiley.com/doi/full/10.1111/ajps.12627) and the [Appendix](https://onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1111%2Fajps.12627&file=ajps12627-sup-0001-Appendix.pdf)


Optional reading:

-   [Design lifecycle](https://book.declaredesign.org/lifecycle) @blair2023research p 319 - 358
-   [DD, Ch 2](https://book.declaredesign.org/introduction/what-is-a-research-design.html) introduces ideas at a high level
-   [@alvarez2018research](https://www.researchgate.net/publication/324240650_Research_Replication_Practical_Considerations/link/5ca3688b458515f7851d7b2e/download?)
-   [@humphreys2013fishing](https://macartan.github.io/assets/pdf/papers/2013_fishing.pdf) on fishing and registration
-   @kurzban2001can
-   @mousa2020building
-   @dietrichSeeingRacialAvoidance2023a

## Causality

Aims:

-   Understand the potential outcomes framework
-   Understand identification challenges and the role of randomization
-   Understand varieties of estimands

**To do before class**: 1. the readings; 2. collectively complete a pre-analysis plan and ethics application for Experiment 1

Required:

-   [Holland](https://www.jstor.org/stable/2289064) is a beautiful classic reading on the fundamental problem of causal inference
-   [DD, Ch 7](https://book.declaredesign.org/declaration-diagnosis-redesign/defining-inquiry.html)

Optional:

-   [II Ch 2](https://integrated-inferences.github.io/book/02-causal-models.html), goes over both potential outcomes and DAGs
-   [@keele2015statistics](https://hrr.w.uib.no/files/2019/01/Keele_2015_causal.pdf) gives a good high level overview of many key ideas in this course
-   [@imai2008misunderstandings](https://gking.harvard.edu/files/matchse.pdf)
-   [@hernanwhatif](https://miguelhernan.org/whatifbook) (Section 3.1, 7.2, 8.4, 10.1)
-   [@hernanwhatif](https://miguelhernan.org/whatifbook) Ch 6
-   [II Ch 4](https://macartan.github.io/integrated_inferences/HJC4.html)
-   @lundberg2021your

## Answer strategies

Aims:

-   Understand alternative strategies to estimating treatment effects
-   Understand design based strategies for estimating standard errors and $p$-values
-   Understand confidence intervals
-   Understand the role of covariates in experimental design

**To do before class**: 1. Collective presentation on the implementation and findings from Experiment 1

Optional:

-   [@freedman2008regression](https://www.stat.berkeley.edu/~census/neyregr.pdf) helps make connections between design based inference and regression
-   [@bind2020possible](https://scholar.harvard.edu/files/marie-abele/files/pnas.pdf) on RI
-   [@lin2012agnostic](https://arxiv.org/pdf/1208.2301.pdf) relieves some worries you might have after reading @freedman2008regression
-   [@chang2024exact](https://onlinelibrary.wiley.com/doi/full/10.3982/ECTA20289) follows up on @lin2012agnostic.

## Experimental Design and Evaluation

Aims:

-   Understand alternative assignment strategies
-   Understand statistical power
-   Understand how to assess alternative diagnosands

**To do before class**: 1. The readings 2. Share initial design for for survey based experiment.

Required:

-   [DD, Ch 7](https://book.declaredesign.org/declaration-diagnosis-redesign/diagnosing-designs.html)
-   [DD, Ch 8](https://book.declaredesign.org/declaration-diagnosis-redesign/crafting-data-strategy.html) on data strategies
-   [DD, Ch 9](https://book.declaredesign.org/declaration-diagnosis-redesign/choosing-answer-strategy.html) on answer strategies
-   [DD, Ch 13](https://book.declaredesign.org/declaration-diagnosis-redesign/declaration-in-code.html) gives more practical guidance on getting started

## Topics

Aims: \* Topics TBD

**To do before class**: 1. The readings 2. Round-robin report on design performance. Teams receive a class-mate's design, declare a model that produces possible data from the design (without consulting the authors design declaration, if any) and applies the original authors answer srtategy, reporting performance.

Required reading (TBD):

### downstream experimentation

-   @coppock2016voting

### spillover experiments

-   @sinclair2012detecting
-   [@aronow2017estimating](https://arxiv.org/abs/1305.6156) on spillovers

### patient preference trials

-   @de2019persuading

### survey experiments

-   @hainmueller2014causal

### mediation experiments

-   [@imai2010general](https://imai.fas.harvard.edu/research/files/BaronKenny.pdf) on mediation

\newpage

# Prerequisites and tools

You should already have background in statistics up to the point of feeling comfortable with regression.

In addition you should know some `R`. Really, the more you can invest on getting on top of `R` before the class the better.

Before the first class please make sure your `R` is up-to-date and that you are working in `R studio`. Then make sure you have the following packages installed.

```{r, eval = FALSE}
pacman::p_load(
  rstan, 
  dagitty, 
  DeclareDesign, 
  CausalQueries,
  tidyverse,
  knitr
  )
```

Resources:

-   Resources for learning `R`: http://www.r-bloggers.com/how-to-learn-r-2/
-   Hadley Wickham's [R for Data Science](https://r4ds.hadley.nz/)
-   [Grant McDermott on data science](https://grantmcdermott.com/ds4e)

## Writing with `qmd`

Please plan to do drafting in `Quarto`. This is a simple markup language that lets you integrate writing and coding. This document is written in `Quarto` and the slides will be also.

The key thing is that you can insert code chunks like this.

```{r}
# Define a random number
x <- rnorm(1)
```

Code like this is run as the document compiles, and results can be accessed as needed, like this: we just sampled the random number $x =$ `r x`.

-   See [@wickham2023r](https://r4ds.hadley.nz/) on [Workflow](https://r4ds.hadley.nz/workflow-scripts) and [Quarto](https://r4ds.hadley.nz/quarto)

I recommend using `Rstudio` as an editor. More information here: https://quarto.org/docs/get-started/hello/rstudio.html

## AI

AI can be a great aid to help understand this material and also to develop code. However it should only be used as an aid: it makes many many mistakes; you remain responsible for your code and more importantly for your comprehension.

# References
